Activity: Selecting Library Books
========================================================

```{r echo=FALSE,error=FALSE,message=FALSE,warning=FALSE,results="hide"}
require(mosaic, quietly=TRUE)
```

This activity shows how bias in sampling can arise in subtle ways.

Students will pick library books from a set of shelves and then calculation the page count for their sample and for the population.

## Set up

* Load the `mosaic` package.
* Give this command:
```{r}
fetchData("simulate.r")
```
At this point, the software is loaded and the `select.books()` function can be run --- instructions below.

## The Activity

Each student selects 10 books and gets back a data from including information on the complete population of books as well as which ones were selected. The measurement on each book is the page count.
```{r eval=FALSE}
books = select.books(10)
```

```{r echo=FALSE,results="hide"}
# Trick the system into running without a mouse input
books = select.books(0)
inds = sample( 1:nrow(books), size=10)
books$selected = as.character(books$selected)
books$selected[inds] = "Yes"
books$selected = as.factor(books$selected)
```

Compare the page count for the selected books to the page count of the unselected books.

```{r}
bwplot( pages ~ selected, data=books)
```

Ask around the class: Do you see any difference between the selected books and the unselected ones.

## A Formal Random Process

We use computers to generate random numbers.  One way to do this is to create a list of books and choose some at random to sample.  For instance, if we had the titles in a list, we could sample from those.  To illustrate, here's a random sample of cases from the KidsFeet data (which has names)
```{r}
sample(KidsFeet, size=3)
```

To see what's different about a formal random sample, the students can simulate one by shuffling the `selected` variable:
```{r}
bwplot( pages ~ shuffle(selected), data=books)
```

We'll use operators like `sample` and `shuffle()` throughout the course.

## In the Future

We'll return to this setting later in the course to examine issues relating to sample size, power, and the alternative hypothesis.

