An Example Problem
========================================================

```{r echo=FALSE}
..QuestionCount.. <- 0
..TextCount.. <- 0
NextQuestionNumber <- function(){
  ..QuestionCount.. <<- ..QuestionCount..+1
  return(..QuestionCount..)
}
NextTextNumber <- function(){
  ..TextCount.. <<- ..TextCount..+1
  return(..TextCount..)
}
Question <- function(name=NextQuestionNumber(),...){
  dots <- list(...)
  nms <- names(dots)
  res <- paste("<select name='in",name,
               "'><option value='NA' selected>UNANSWERED</option>",
               sep="")
  for (k in 1:length(dots)) {
    opt <- paste("<option value='",dots[[k]],"'>",nms[k],"</option>",sep="")
    res <- paste(res, opt)
  }
  res <- paste(res, 
               paste("</select><span id='out",name,
                     "' class='shiny-html-output'> </span>",sep=""))
  return(res)
}
TrueOrFalse <- function(name=NextQuestionNumber(),ans=TRUE) {
  Question(name=name,"TRUE"=ans,"FALSE"=!ans)
}

TextEntry <- function(name=NextTextNumber(),default=""){
  res <- paste("<textarea cols='60' rows='5' name='text",
               name,"'>",default,"</textarea>",sep="")
  res <- paste(res, paste("<span id='tout",name,
                          "' class='shiny-html-output'></span>",sep=""))
  return(res)
}
```
### Text questions

Write an essay about something:

`r I(TextEntry())`


### Here's one of my new questions: 

Which of the following is a true statement? `r I(Question(Argentina="No",Russia="Wrong-o",Canada="You betcha!"))`
* Argentina is a country near China?
* Russia is a country near Bolivia?
* Canada is a country near the US?

### And a True/False Question

* The rain in Spain falls mainly on the plain. `r I(TrueOrFalse(ans=TRUE))`


```{r}
plot(1:20, pch=1:20)
```