\documentclass[12pt]{article}
\pagestyle{empty}

\begin{document}
\vspace*{-1in}
\centerline{\Large \bfseries Small Group Discussion Questions}

\bigskip

\centerline{\large \sf Epidemiology}

\bigskip

\begin{enumerate}
\item Suppose you are doing research on a new antibiotic. it's not likely, but some people might be 
allergic and die from it. 

Can you ethically do the research? 
\item You want to do research on an alternative psychological treatment 
involving complete isolation in traditional sweat baths after fasting. The idea is that coming close to death helps increase spirituality.  

Can you ethically do the research?

\item Suppose you are testing a new, very inexpensive drug 
that likely won't perform as well as the current 
standard drug that's very expensive? 

Can you ethically do the research?

\item How would you do ethical research into the best way 
to revive someone who has hypothermia? You'll need to submerge people in cold water to create the hypothermia.

Can you ethicially do this research?

\item Suppose you are in the midst of a Randomized Clinical Trial.  A patient decides to stop involvement because he's not interested any more. To what extent are you allowed to pursuade him of the importance of the study?  

Now suppose that the patient wants an additional payment to continue his participation.  Should you provide it?

\item There have been many studies "intercessional prayer" on patient outcomes.  A typical study design is to have third parties  (say, members of a church congregation) pray for some patients in the ICU, while other patients will not be prayed for by the third parties (the control).  The investigators randomly assign ICU patients to prayer or control.

Suppose that you are a journal editor.  The researchers have not gotten informed consent from the ICU patients.  The researchers argue that it wouldn't be polite to ask very sick people if they want to be prayed for, and then possibly deny them the prayer.  It might even be stressful to the patient and his or her family to be approached with this request.

Can you ethically publish the results of the research?
\end{enumerate}

\end{document}