\begin{multicols}{2}
\paragraph{Question 1}

One way to measure the biologically relevant exposure to ionizing
radiation is in sieverts (sV). 

\begin{itemize}
\item An accumulated exposure to 1 sV over months increases the lifetime incidence of cancer by
  about 5\%.
\item The smallest accumulated radiation dose linked to any increased cancer risk is
  0.1 sV.
\item Intense short-term dose can lead to radiation poisoning, which
  is potentially fatal.
\end{itemize}

Based on this information, give your estimation of the number of sV
associated with each of the following activities:

\begin{enumerate}[(a)]
\item Normal yearly dose.
\item Extra dose from a cross-US airplane flight.
\item Extra dose from the potassium in eating one banana.
\item Extra dose from a CT scan of the head.
\item Extra dose from a mammogram.
\item Average annual extra dose from living within 10 miles of the Three-Mile Island
  nuclear plant since the 1979 accident. 
\item Extra dose to a person in Tokyo following the Fukushima accident.
\item EPA dose limit for emergency workers in lifesaving operations.
\item Dose received by most severely exposed Fukushima plant workers.
\item Dose causing symptoms of radiation poisoning.
\item Acutely fatal radiation-poisoning dose (death within days, not from cancer)
\end{enumerate}

\paragraph{Question 2}

\begin{enumerate}[(a)]
\item Is the Mississippi river shorter or longer than \mississippi\ miles?
\item How long is the Mississippi river?
\end{enumerate}


\paragraph{Question 3}

Linda is 31 years old, single, outspoken, and very bright. She
majored in philosophy. As a student, she was deeply concerned with issues of
discrimination and social justice, and also participated in antinuclear rallies.
Rank the following possibilities from most likely to least likely, for instance “A
B C D E”:
\begin{enumerate}[(A)]
\item Linda is a teacher.
\item Linda works in a bookstore and takes yoga classes.
\item Linda is a bank teller.
\item Linda sells insurance.
\item Linda is a bank teller and is active in the feminist movement.
\end{enumerate}

\end{multicols}
