Monday April 29, 2013 Stats 155 Class Notes 
================================

```{r name="setup", child="notes-setup.Rmd"}
# boilerplate
```
$latex \newcommand{\ensuremath}[1]{#1}
\newcommand{\pathway}[1]{\hbox{#1}}
\newcommand{\causes}{\ensuremath{\Rightarrow}}
\newcommand{\causedBy}{\ensuremath{\Leftarrow}}
\newcommand{\correlatedWith}{\ensuremath{\nLeftrightarrow}}$



```{r tikz-ex, echo=FALSE,eval=FALSE,engine = "tikz", fig.cap = "Funky tikz", fig.ext = 'png', cache=TRUE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=5cm, auto,>=latex', thick, scale = 2.0]
\node (P) {$P$};
\node (B) [right of=P] {$B$};
\node (A) [below of=P] {$A$};
\node (C) [below of=B] {$C$};
\node (P1) [node distance=1.4cm, left of=P, above of=P] {$\hat{P}$};
\draw[->] (P) to node {$f$} (B);
\draw[->] (P) to node [swap] {$g$} (A);
\draw[->] (A) to node [swap] {$f$} (C);
\draw[->] (B) to node {$g$} (C);
\draw[->, bend right] (P1) to node [swap] {$\hat{g}$} (A);
\draw[->, bend left] (P1) to node {$\hat{f}$} (B);
\draw[->, dashed] (P1) to node {$k$} (P);
\end{tikzpicture}
```

#### Confounders as a diagram

```{r confounder, engine = "tikz", fig.cap = "Funky tikz", fig.ext = 'png', cache=TRUE, echo=FALSE}
\usetikzlibrary{arrows,shapes,snakes}
\begin{tikzpicture}[node distance=5cm, auto,>=latex', thick, scale = 2.0]
\node[rectangle,fill=blue!10] (V) {\LARGE Variable};
\node[rectangle,fill=blue!10] (Out) at ([shift={(3,0)}] V) {\LARGE Outcome};
\node[ellipse,fill=red!10] (A) at ([shift={(1.5,-1.5)}] V) {\LARGE Confounder};
\draw[->,ultra thick] (V) to (Out);
\draw[<->,ultra thick,bend right] (V) to (A);
\draw[<->,ultra thick,bend right] (A) to (Out);
\end{tikzpicture}
```

#### SAT versus per-student spending

Draw the causal diagram

```{r sat, engine = "tikz", fig.cap = "Funky tikz", fig.ext = 'png', cache=TRUE, echo=FALSE}
\usetikzlibrary{arrows,shapes,snakes}
\begin{tikzpicture}[node distance=5cm, auto,>=latex', thick, scale = 2.0]
\node[rectangle,fill=blue!10] (Spend) {\LARGE Spending};
\node[rectangle,fill=blue!10] (SAT) at ([shift={(3,0)}] Spend) {\LARGE SAT};
\node[ellipse,fill=red!10] (Focus) at ([shift={(-.5,-2)}] Spend) {\LARGE \parbox{3cm}{\raggedright Focus on Education}};
\node[rectangle,fill=blue!10] (Frac) at ([shift={(.5,-2)}] SAT) {\LARGE \parbox{3cm}{\raggedleft Fraction taking SAT}};
\draw[->,ultra thick] (Spend) to (SAT);
\draw[->,ultra thick] (Focus) to (Frac);
\draw[<->,ultra thick,bend left] (Focus) to (Spend);
\draw[->,ultra thick] (Frac) to (SAT);
\end{tikzpicture}
```


<pre>
Spending  ---------->      SAT    <-------|
      |                                   |
Focus on Educ. ---> fraction taking SAT  -|
</pre>

#### Campaign Spending and the back-door network

Research in political science shows that higher spending in campaigns is related to a lower vote for the incumbent.  Yet it's common sense that higher spending improves things for the candidate; that's why they do it.

<pre>
  Polls <-----    Popularity ---> vote outcome
    |                                 ^
    v                                 |
  Spending ---------------------------
</pre>

### How to block a back-door?

We've done it by including the covariate in the model.  But this is too crude an answer.

### Example: Election Spending

```{r}
fetchData("simulate.r")
campaign.spending
equations(campaign.spending)
```
You can see from the equations how spending is related to vote: it increases it.  Let's look at what a statistical model has to say
```{r}
d = run.sim(campaign.spending, 435 ) # number of congressment
summary( lm( vote ~ spending, data=d))
```
The problem: a back door pathway from spending to vote via popularity.  Block it by including a node on the pathway as a covariate.

```{r}
summary( lm( vote ~ spending + polls, data=d))
```


