February 18, 2013 Class Notes
========================================================

```{r include=FALSE}
require(mosaic)
``` 

### Back to RMarkdown

Several people had problems compiling their documents.  These were caused by ill-formed commands earlier in the document.

Use the "Chunks" menu to run individual chunks by copying them to your console.

Write up a short document about partitioning variance and post it to Moodle.

### Notation

In the `lm()` program (and `mm()` and `glm()` programs), we write the response variable to the left of the `~` and the model terms, separated by the `+` notation, on the right side.  Don't confuse `+` with mathematical addition.  It really is just a separator for the various terms we are going to be including in the linear combination that will constitute the model.

* The `1` means "intercept".  As it happens, the software will automatically include the intercept even if you don't say `1`.  If you want to exclude the intercept, you use a special notation: `-1`.

Example of simple models with and without the intercept term.
```{r}
lm( height ~ 1 + father, data=Galton )
lm( height ~ father, data=Galton ) # same thing as previous
lm( height ~ father - 1, data=Galton ) # leave out intercept
```

* The name of the variable, e.g. `father` means that variable as a main effect.

* An interaction term is introduced with a `:` between variable names.
```{r}
lm( height ~ father + mother + father:mother, data=Galton )
```

* Since the polynomial modeling heuristic includes main effects by default, there is a special notation to signal that we want both the main effects and the interaction:
```{r}
lm( height ~ father*mother, data=Galton )
```

* Transformation terms: apply the function to the variable. Example:
```{r}
lm( height ~ cos(father), data=Galton ) # silly model!
swim = fetchData("swim100m.csv")
swimmod = lm( time ~ sex * exp( -year/50 ), data=swim) 
xyplot( time + fitted(swimmod) ~ year, data=swim)
```

(N.B.: For instructors ... Beware that the above exponential model just barely makes it within the floating-point arithmetic capabilities of the computer.  Applying an exponential to numbers arguments like $latex -20$ can be perilous.  Better to have rescaled year to, say, start at 1900.)

* Quadratic terms are fairly common, and so there is a nice way to specify that you want both the main effect and higher-order polynomial terms:
```{r}
swimmod2 = lm( time ~ sex*poly(year,2), data=swim)
xyplot( time + fitted(swimmod2) ~ year, data=swim)
```

N.B. For instructors ... You may be tempted to write polynomial terms directly, for instance
```{r}
swimmod3 = lm( time ~ sex*( year + I(year^2)), data=swim)
xyplot( time + fitted(swimmod3) ~ year, data=swim)
``` 
Note the use of `I()` to instruct R to treat arithmetical symbols in the numerical way.  Also, note that `poly()` cleverly does a shift into orthogonal polynomials to preserve numerical stability.  As such, the specific coefficients in `swimmod2` and `swimmod3` will be different, even though the model formulas amount to the same thing.
```{r}
swimmod2
swimmod3
```
In order to deal with the possible double meaning of symbols like `+`, you can use `I()` to signal to R to treat the mathematical operations literally.

Introducing $latex R^2$ Briefly
--------------------------
Since we're trying to account for variation in the response variable, it's helpful to have a way to quantify how much variation we have accounted for.  The quantity called "R-squared" --- $latex R^2$, officially "the coefficient of determination" although hardly anyone uses that name --- is a standard way to do this.  Briefly, $latex R^2$ is the fraction of variance in the response variable that is "captured" by the model.  It is always between 0 (didn't capture any) and 1 (got it all!).



### Activity
Visual choice of model terms

Instructions are [given here](http://dl.dropbox.com/u/5098197/Exercises/HTML/ISM-Selecting-Model-Terms/ISM-Selecting-Model-Terms-A.html?access=ISM3)  (These have to be reformatted as a handout and as an instructor's guide.)

The only substantive differences is in the way the software is loaded:
```{r}
fetchData("mLM.R")
```

The program itself is called `mLM()` (not `lm.select.terms()`)

Model Formulas and Coefficients
-------------------------------

Fit some models and work students through the calculations.  The `makeFun()` function can help in this.  

A suggested style for a calculation:
```{r}
mod = lm( wage ~ sex + age, data=CPS85 )
coef(mod)
```
What is the wage for a female of age 40?  A male of age 30?  Work through the arithmetic "by hand".
```{r}
4.65 + 0*2.274 + 40*0.0852 # for the 40-year old female
4.65 + 1*2.274 + 30*0.0852 # for the 30-year old male
```

You can also turn the model information into a mathematical function:
```{r}
f = makeFun(mod)
f(sex="F", age=40)
f(sex="M", age=30)
```

Other settings:
* Natural gas usage by average temperature in `utilities.csv`
* ... more ...

## Form Groups for Car Project

Outline the [used car project](https://dl.dropbox.com/u/5098197/Exercises/HTML/SM2-Chapter-7-Problems/SM2-Chapter-7-Problems-A.html?access=ISMf12#723) and show some data from <http://www.cars.com>.  Perhaps MINI Coopers.

Get students to form groups right now, oriented around different kinds of cars.  They should make a Google Spreadsheet and share it with all the other members of their group.

