Wednesday April 3 Class Notes
================================

```{r name="setup", child="notes-setup.Rmd"}
# boilerplate
```



### Review of Vocabulary

A graphical way of thinking about hypothesis testing.

App for playing with Significance and Power

```{r eval=FALSE}
fetchData("mHypTest.R")
mHypTest() # by default, a coefficient
```


### Power

How did I know that I could reject the Null in the shuffling problem?  I did a little simulation.
```{r}
mysim <- function(n=15) {
   days = resample(1:31,size=n)
   nums = ceiling( runif(n,min=0,max=days))
   mod = lm(nums~days)
   list(r2=r.squared(mod), p=summary(mod)$coef[2,4])
}
s15 = do(1000)*mysim(24) # typical R^2 is about 0.4
mean(~r2, data=s15)
tally( ~ p<0.05, data=s15,format='proportion')
```

### Hypothesis tests on individual coefficients

The p-value on an individual coefficient.

Simulate life on Planet Null by shuffling the variables involved in the coefficient.

What's the distribution of p-values on Planet Null?

```{r}
summary(lm(width ~ sex + length, data=KidsFeet))$coef
summary(lm(width ~ shuffle(sex) + length, data=KidsFeet))$coef
summary(lm(width ~ shuffle(sex) + length, data=KidsFeet))$coef[2,4]
s = do(1000)*summary(lm(width ~ shuffle(sex) + length, data=KidsFeet))$coef[2,4]
densityplot(~result,data=s)
```


### R^2 as a test statistic

What's the distribution of R^2 under the Null Hypothesis?

What do we mean by the Null Hypothesis?
* All explanatory variables are unrelated to the response.
* Just one explanatory variable is related?
* Some subset of explanatory variables are related?

We're going to study the "Whole Model" Null Hypothesis.  

Show distribution of R^2.

### It's nice to have some theory. 

What's the typical value of R^2 if you make up $m-1$ random vectors to explain $n$ data points.  (We'll always include an intercept in the model in addition to the random vectors.)

```{r}
kids = fetchData("KidsFeet")
r.squared( lm( width ~ rand(10), data=kids))
s = do(1000)*r.squared( lm( width ~ rand(10), data=kids))
densityplot(~result, data=s)
```

* Do it for no random vectors.
* Do it for 40 random vectors.  R^2 is always 1
* What's the smallest $m-1$ can be to get an R^2 of 1?
* Do it for 1, 2, 3, 4, random vectors, 10, 20, 30 random vectors.  What's the pattern?

Now do it for the `CPS85` data.  Tell me how many random vectors you would need to get a typical R^2 of 0.2.  Then simulate this and confirm your answer.

Do `mHypTest(TRUE)` setting the "effect size" to about 0.4
Translation to F.

```{r}
s = do(1000)*r.squared( lm( width ~ rand(10), data=kids))
s = transform(s, F=(result/10)/((1-result)/(38-10)))
densityplot(~F, data=s)
plotFun(df(x,10,28)~x, add=TRUE, col="red")
```