Friday 26 April
===========================

## Survey processing.

Example of survey processing commands [here](http://rpubs.com/dtkaplan/m155-processing-survey-data) and the [raw template](https://github.com/dtkaplan/Statistical-Modeling-A-Fresh-Approach/blob/gh-pages/Activities/SurveyProcessing.Rmd).

## Introduction to Causal Reasoning

### Myopia

* [Myopia article from Nature](http://dl.dropbox.com/u/5098197/ISM/myopia.pdf).  Note the extensive scientific jargon and the dismissing of the lack of causal evidence: "Although it does not establish a causal link, the statistical strength of the association of night-time light exposure and childhood myopia does suggest that the absence of a daily period of darkness during early childhood is a potential precipitating factor in the development of myopia."

It turns out that the parent's myopia is the disposing factor.  Shortsighted parents leave the lights on at night and their kids are shortsighted for genetic reasons.  A CNN news report is [here](http://www.cnn.com/HEALTH/9905/12/children.lights/index.htm). And here is the [abstract of the rebuttal](http://www.nature.com/nature/journal/v404/n6774/full/404143a0.html).


```{r tikz-ex, echo=FALSE,eval=FALSE,engine = "tikz", fig.cap = "Funky tikz", fig.ext = 'png', cache=TRUE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=5cm, auto,>=latex', thick, scale = 2.0]
\node (P) {$P$};
\node (B) [right of=P] {$B$};
\node (A) [below of=P] {$A$};
\node (C) [below of=B] {$C$};
\node (P1) [node distance=1.4cm, left of=P, above of=P] {$\hat{P}$};
\draw[->] (P) to node {$f$} (B);
\draw[->] (P) to node [swap] {$g$} (A);
\draw[->] (A) to node [swap] {$f$} (C);
\draw[->] (B) to node {$g$} (C);
\draw[->, bend right] (P1) to node [swap] {$\hat{g}$} (A);
\draw[->, bend left] (P1) to node {$\hat{f}$} (B);
\draw[->, dashed] (P1) to node {$k$} (P);
\end{tikzpicture}
```

#### Confounders as a diagram

```{r confounder, engine = "tikz", fig.cap = "Funky tikz", fig.ext = 'png', cache=TRUE, echo=FALSE}
\usetikzlibrary{arrows,shapes,snakes}
\begin{tikzpicture}[node distance=5cm, auto,>=latex', thick, scale = 2.0]
\node[rectangle,fill=blue!10] (V) {\LARGE Variable};
\node[rectangle,fill=blue!10] (Out) at ([shift={(3,0)}] V) {\LARGE Outcome};
\node[ellipse,fill=red!10] (A) at ([shift={(1.5,-1.5)}] V) {\LARGE Confounder};
\draw[->,ultra thick] (V) to (Out);
\draw[<->,ultra thick,bend right] (V) to (A);
\draw[<->,ultra thick,bend right] (A) to (Out);
\end{tikzpicture}
```

#### SAT versus per-student spending

Draw the causal diagram

```{r sat, engine = "tikz", fig.cap = "Funky tikz", fig.ext = 'png', cache=TRUE, echo=FALSE}
\usetikzlibrary{arrows,shapes,snakes}
\begin{tikzpicture}[node distance=5cm, auto,>=latex', thick, scale = 2.0]
\node[rectangle,fill=blue!10] (Spend) {\LARGE Spending};
\node[rectangle,fill=blue!10] (SAT) at ([shift={(3,0)}] Spend) {\LARGE SAT};
\node[ellipse,fill=red!10] (Focus) at ([shift={(-.5,-2)}] Spend) {\LARGE \parbox{3cm}{\raggedright Focus on Education}};
\node[rectangle,fill=blue!10] (Frac) at ([shift={(.5,-2)}] SAT) {\LARGE \parbox{3cm}{\raggedleft Fraction taking SAT}};
\draw[->,ultra thick] (Spend) to (SAT);
\draw[->,ultra thick] (Focus) to (Frac);
\draw[<->,ultra thick,bend left] (Focus) to (Spend);
\draw[->,ultra thick] (Frac) to (SAT);
\end{tikzpicture}
```


<pre>
Spending  ---------->      SAT    <-------|
      |                                   |
Focus on Educ. ---> fraction taking SAT  -|
</pre>

#### Campaign Spending and the back-door network

Research in political science shows that higher spending in campaigns is related to a lower vote for the incumbent.  Yet it's common sense that higher spending improves things for the candidate; that's why they do it.

<pre>
  Polls <-----    Popularity ---> vote outcome
    |                                 ^
    v                                 |
  Spending ---------------------------
</pre>

### How to block a back-door?

We've done it by including the covariate in the model.  But this is too crude an answer.

### Example: Election Spending

```{r}
fetchData("simulate.r")
campaign.spending
equations(campaign.spending)
```
You can see from the equations how spending is related to vote: it increases it.  Let's look at what a statistical model has to say
```{r}
d = run.sim(campaign.spending, 435 ) # number of congressment
summary( lm( vote ~ spending, data=d))
```
The problem: a back door pathway from spending to vote via popularity.  Block it by including a node on the pathway as a covariate.

```{r}
summary( lm( vote ~ spending + polls, data=d))
```
