Friday April 12, 2013 Stats 155 Class Notes 
==================

```{r name="setup", child="notes-setup.Rmd"}
# boilerplate
```

### Graphical depictions of ANOVA

* R^2 versus m graph
* Variable-space graph.  How far did the original term get you?  How much further does the next term get you?

Constrast with variable-space depiction of regression report.  Order doesn't matter.

### Review

* Two different standard reports for models
    * regression report
    * ANOVA report
* Regression puts each variable in the context of all the others
* ANOVA puts each variable into the context of
    * What gets added onto what the previous variables can explain
    * Cleaning up residuals with the following variables.


## Commonly Used Tests

But not at all the general use for ANOVA

### One-way ANOVA (so called)


An attempt to be honest with p-values and avoid the problem of multiple comparisons.

Several levels of a categorical variable.  Null hypothesis is generally phrased as "the levels are all the same."

Compare to t-test: two levels.

Professors: many levels.

### Two categorical variables

Typical central question: Is there an interaction between the variables?

Example: Sector and sex

```{r}
mod = lm( wage ~ sector*sex, data=CPS85)
anova(mod)
```
Have we detected an interaction?

#### Digression: Log prices, indices, and wages

Prices are relative.  An indication of this is the almost universal use of percentages to describe inflation, wage increases, etc.  For example, an often quoted number is that women earn approximately 72 cents for each dollar earned by a man.  

The naive way to find this number, which is in fact the way it is found, is to divide the average wage of women by the average wage of men, e.g.
```{r}
mean( wage ~ sex, data=CPS85)
7.88/9.99
```
Close to the quoted number in this data set from the 1980s.

A better way is to work with log wages, find the contribution from sex, and then convert that back into a multiplier.  That let's us adjust for various other factors.  Here's the basic calculation, done in log-wage style:
```{r}
lm( log(wage) ~ sex, data=CPS85 )
exp(-0.2312)
```

Now we can include covariates:
```{r}
lm( log(wage) ~ sex + sector + exper + educ, data=CPS85 )
exp(-0.2197)
```

Hardly any difference.  But maybe the model should be more complicated. 

#### Activity

Use ANOVA and log wages to see if interactions should be included in the model.  Example:
```{r}
mod = lm(log(wage) ~ sector*sex, data=CPS85)
anova(mod)
```

####  Student Activity
Question: Is there an interaction between age and mileage in the used car data?  Does it show up if we look at log prices?
```{r}
cars = fetchData("used-hondas.csv")
anova(lm(log(Price)~Age*Mileage, data=cars))
```


### Example: Do professors vary in how they grade? Revisited

One place where F shines is when we want to look at many explanatory vectors collectively.

In the last class, we looked at professor-wise gradepoint averages, with an eye to the question of whether some professors are easy grading.  We used as a test statistic, the model coefficient for each professor, and ran into the question of multiple comparisons.  

Now let's return to the question using analysis of variance.

```{r}
grades = fetchData("grades.csv")
courses = fetchData("courses.csv")
g2n = fetchData("grade-to-number.csv")
all = merge( grades, courses)
all = merge(all, g2n) # a data set of every grade given, etc.
```

Suppose, instead of being concerned about individual professors, we were interested in the professorate as a whole: do they grade in a consistent way, where "consistent" means, "draw grades from a common pool."  This test can be done easily.  Build the model and see if the explanatory variable accounts for more than is likely to arise from chance:
```{r}
mod1 = lm( gradepoint ~ iid, data=all )
r.squared(mod1)
```

The regression report actually gives a p-value for this r.squared.  It's not any different than we would get by travelling to Planet Null:  randomizing `iid` and seeing what is the distribution of R^2 on Planet Null.

Another way to summarize the model is with an ANOVA report:
```{r}
anova(mod1)
```

### Breaking up the Variance into Parts

Of course, it's not fair to credit professors for variation in grades that is really due to the students.  So we want to divide up the variation into that due to the students and that due to the professors.  ANOVA let's you do this:
```{r}
mod2 = lm( gradepoint ~ sid + iid, data=all )
anova(mod2)
```

Interestingly, the result depends on the order in which you put the model terms, even though the model values do not at all depend on this.
```{r}
mod3 = lm( gradepoint ~ iid + sid, data=all )
anova(mod3)
sum( (fitted(mod2) - fitted(mod3))^2) # model values are the same
```

Eating Up the Variance
----------------

The F statistic compares the "credit" earned by a model term to the mean square residual, which can be interpreted as the credit that would be earned by a junky random term.

Fit a model and add in some random terms.  Show that the F for the random terms is about 1 and that the mean square of the residual is hardly changed by the random terms.

```{r}
mod0 = lm(wage ~ sector + sex + exper, data=CPS85)
anova(mod0)
```
The mean square residual is about 20.

Now throw in some junk:
```{r}
mod10 = lm( wage ~ sector + sex + exper + rand(10), data=CPS85)
anova(mod10)
```

But what if a term eats more variance than a junky term.  That term makes it easier for the other terms to show significance.

EXAMPLE:  Difference in age between husband and wife in couples getting married.

Ask: Who is older in a married couple, the man or the woman?  By how much?  

Let's see if the data support this:
```{r}
m = fetchData("marriage.csv")
mod0 = mm( Age ~ Person, data=m )
mod0
confint(mod0)
mod1 = lm( Age ~ Person, data=m )
summary(mod1)
```
The point estimate is about right, but the margin of error is so large that we can't take this estimate very seriously.  The p-value is so large that we can't reject the null that there is no relationship between `Person` and age.

Try adding in some other variables, astrological sign, years of education, etc. and show that this doesn't help much.

Finally, add in the `BookpageID` variable.  
```{r}
mod2 = lm( Age ~ Person + BookpageID, data=m )
anova(mod2)
head(confint(mod2))
```

This gives an individual ID to each marriage.  Putting this in the model effectively holds the couple constant when considering the `Person`.  In terms of ANOVA, `BookpageID` is eating up lots and lots of variance.

### ACTIVITY

Consider the `SAT` data and `sat` as the response variable.  Make models with `frac`, `salary`, `expend` and `salary` as the explanatory variables.  Try different orders for the explanatory variables.  Try both the ANOVA and the regression reports.  Come up with a theory of what's happening.


