Monday April 8, 2013 Stats 155 Class Notes 
==================

```{r name="setup", child="notes-setup.Rmd"}
# boilerplate
```


## Overview


#### An Example

An article about organic food diet and health of young children in the [British Journal of Nutrition](https://dl.dropbox.com/u/5098197/ISM/OrganicFoods-14519.pdf).

Table 3 shows odds ratios for five different foods for both moderately organic and strictly organic diets.  10 comparisons, one of which is significant at $p=0.02$.  Doing a Bonferroni adjustment suggests an appropriate threshold is 0.005.

### Interpreting p-values

### Activity

Print out [list of 20 p-values](../Epidemiology/Notes/Activities/conf-interval-tables.pdf). Ask who has got the first one < 0.05? Who has got any of them? What's the smallest p-value in the class?

Simulation activity:
```{r echo=-1}
set.seed(123)
fetchData("simulate.r")
d = run.sim( bogus.groups, ngroups=5, 10)
mean(val~group, data=d)
dd = subset( d, group %in% c("D", "C"), drop=TRUE ) # or whatever are the furthest-apart groups
summary(lm(val~group, data=dd))
```

Carry out a hypothesis test on a subset that is just the best and worst groups. How often do people get a p-value less than 0.05?

Problem 14.04.

Publication bias.  Filing cabinet problem.


### mLM and reading R^2 as more terms change

Show how terms increase R^2, which gives the fraction of distance to the response variable.

Introduce the idea of the log to handle the interaction between sector and sex, why the dollar amount of difference between the sectors should differ between the sexes.


### The Structure of the ANOVA table

Starting model: `A ~1`

The basic information on each additional term:
* Increment in the sum of squares of response (difference from last model)
* Degrees of freedom (increase in number of non-redundant coefficients)

Final model: `A ~ ... + rand(big)`


### Two categorical variables

Example: Sector and sex

```{r}
mod = lm( wage ~ sector*sex, data=CPS85)
anova(mod)
```
Have we detected an interaction?

#### ACTIVITY

Create the set of nested models and construct the ANOVA table by hand.

#### Digression: Log prices and wages

Prices are relative.  An indication of this is the almost universal use of percentages to describe inflation, wage increases, etc.  For example, an often quoted number is that women earn approximately 72 cents for each dollar earned by a man.  

The naive way to find this number, which is in fact the way it is found, is to divide the average wage of women by the average wage of men, e.g.
```{r}
mean( wage ~ sex, data=CPS85)
7.88/9.99
```
Close to the quoted number in this data set from the 1980s.

A better way is to work with log wages, find the contribution from sex, and then convert that back into a multiplier.  That let's us adjust for various other factors.  Here's the basic calculation, done in log-wage style:
```{r}
lm( log(wage) ~ sex, data=CPS85 )
exp(-0.2312)
```

Now we can include covariates:
```{r}
lm( log(wage) ~ sex + sector + exper + educ, data=CPS85 )
exp(-0.2197)
```

Hardly any difference.  But maybe the model should be more complicated. 

#### Activity

Use ANOVA and log wages to see if interactions should be included in the model.  Example:
```{r}
mod = lm(log(wage) ~ sector*sex, data=CPS85)
anova(mod)
```

####  Student Activity
Question: Is there an interaction between age and mileage in the used car data?  Does it show up if we look at log prices?
```{r}
cars = fetchData("used-hondas.csv")
anova(lm(log(Price)~Age*Mileage, data=cars))
```


### Example: Do professors vary in how they grade? Revisited

One place where F shines is when we want to look at many explanatory vectors collectively.

In the last class, we looked at professor-wise gradepoint averages, with an eye to the question of whether some professors are easy grading.  We used as a test statistic, the model coefficient for each professor, and ran into the question of multiple comparisons.  

Now let's return to the question using analysis of variance.

```{r}
grades = fetchData("grades.csv")
courses = fetchData("courses.csv")
g2n = fetchData("grade-to-number.csv")
all = merge( grades, courses)
all = merge(all, g2n) # a data set of every grade given, etc.
```

Suppose, instead of being concerned about individual professors, we were interested in the professorate as a whole: do they grade in a consistent way, where "consistent" means, "draw grades from a common pool."  This test can be done easily.  Build the model and see if the explanatory variable accounts for more than is likely to arise from chance:
```{r}
mod1 = lm( gradepoint ~ iid, data=all )
r.squared(mod1)
```

The regression report actually gives a p-value for this r.squared.  It's not any different than we would get by travelling to Planet Null:  randomizing `iid` and seeing what is the distribution of R^2 on Planet Null.

Another way to summarize the model is with an ANOVA report:
```{r}
anova(mod1)
```


### Theory of F

Explain R^2 in terms of the graph of (hypothetical) R^2 versus number of junky model vectors.  F is the ratio of segment slopes

Now do it stepwise by finding the sum of squares of the fitted model values in a set of models `~1` and `~1 + iid`
* Sum of squares of the fitted model values
* DF ("degrees of freedom") a count of coefficients
* Residual SS --- sum of squares of the residuals
* Residual DF --- $n$ - DF of the model
* Look at how things change when going from `~1` to `~1 +iid`.
* "Mean square" is SS/DF

A way to think of the F statistic: miles per gallon for the model terms compared to miles per gallon for the 

### Breaking up the Variance into Parts

Of course, it's not fair to credit professors for variation in grades that is really due to the students.  So we want to divide up the variation into that due to the students and that due to the professors.  ANOVA let's you do this:
```{r}
mod2 = lm( gradepoint ~ sid + iid, data=all )
anova(mod2)
```

Interestingly, the result depends on the order in which you put the model terms, even though the model values do not at all depend on this.
```{r}
mod3 = lm( gradepoint ~ iid + sid, data=all )
anova(mod3)
sum( (fitted(mod2) - fitted(mod3))^2) # model values are the same
```

Eating Up the Variance
----------------

The F statistic compares the "credit" earned by a model term to the mean square residual, which can be interpreted as the credit that would be earned by a junky random term.

Fit a model and add in some random terms.  Show that the F for the random terms is about 1 and that the mean square of the residual is hardly changed by the random terms.

```{r}
mod0 = lm(wage ~ sector + sex + exper, data=CPS85)
anova(mod0)
```
The mean square residual is about 20.

Now throw in some junk:
```{r}
mod10 = lm( wage ~ sector + sex + exper + rand(10), data=CPS85)
anova(mod10)
```

But what if a term eats more variance than a junky term.  That term makes it easier for the other terms to show significance.

EXAMPLE:  Difference in age between husband and wife in couples getting married.

Ask: Who is older in a married couple, the man or the woman?  By how much?  

Let's see if the data support this:
```{r}
m = fetchData("marriage.csv")
mod0 = mm( Age ~ Person, data=m )
mod0
confint(mod0)
mod1 = lm( Age ~ Person, data=m )
summary(mod1)
```
The point estimate is about right, but the margin of error is so large that we can't take this estimate very seriously.  The p-value is so large that we can't reject the null that there is no relationship between `Person` and age.

Try adding in some other variables, astrological sign, years of education, etc. and show that this doesn't help much.

Finally, add in the `BookpageID` variable.  
```{r}
mod2 = lm( Age ~ Person + BookpageID, data=m )
anova(mod2)
head(confint(mod2))
```

This gives an individual ID to each marriage.  Putting this in the model effectively holds the couple constant when considering the `Person`.  In terms of ANOVA, `BookpageID` is eating up lots and lots of variance.


### Meta-studies and funnel plots 

* The standard error and the funnel plot: looking at lots of studies and comparing their odds ratios and standard errors: Start at time 9:30 in [this TED video](http://www.ted.com/talks/ben_goldacre_battling_bad_science.html) 
* [Wikipedia](http://en.wikipedia.org/wiki/Funnel_plot)
