February 20, 2013 Class Notes
=============

```{r include=FALSE}
require(mosaic)
options(na.rm=TRUE)
```

Reviewing the car price models.

* [One](http://rpubs.com/drathakr/4539)
```{r}
cars1 = fetchGoogle("https://docs.google.com/spreadsheet/pub?key=0Akkdc3QLDbyAdGxNRDAzcG5aRGJBSVFHa3ljaHVObnc&output=csv")
confint(lm( Price ~ Age, data=cars1))
confint(lm( Price ~ Miles, data=cars1))
confint(lm( Price ~ Age+Miles, data=cars1))
```

* [Two](https://docs.google.com/document/d/1POuNT3nWBnqF7c7CTIEbWnl-MCVK6sVEGcRns8cAIbs/edit)
    [Spreadsheet](https://docs.google.com/spreadsheet/ccc?key=0AnVnVAhrWOVPdDhpaVFiS0RtUTlwVjdyS2xnNnNWSUE#gid=0)
    
Construct models using Zip, Age, Miles and discuss the units.  What happens with interaction terms?  What are the units of those?


```{r}
prius = fetchData("M155/Spring2013/ToyotaPrius.csv") # fetchData("Spring2013/Notes/ToyotaPrius.csv")
lm( Price ~ Age, data=prius)
```
What does the intercept mean?  Price of a hypothetical prius produced in year 0 --- the Roman prius (not to be confused with Priapus).

```{r}
prius = transform(prius, Age=2013-Year)
mod = lm( Price ~ Age+Miles, data=prius)
mod
# make a 2-d plot of this
f1 = makeFun(mod)
plotFun( f(Age=a, Miles=m)~ a&m, 
         a.lim=c(0,10),m.lim=c(0,100000))
lm( Price ~ Age+Miles+Zip, data=prius)
lm( Price ~ Age + Miles*Zip, data=prius)
```

#### Watch out! An outlier in price.

```{r}
mod = lm( Price ~ Age*Miles, data=prius)
f = makeFun(mod)
plotFun( f(Age=a, Miles=m)~ a&m, a.lim=c(0,10),m.lim=c(0,100000))
max( Miles, data=prius)
pp =subset(prius, Miles < 100000)
mod = lm( Price ~ Age*Miles, data=pp)
f = makeFun(mod)
plotFun( f(Age=a, Miles=m)~ a&m, a.lim=c(0,10),m.lim=c(0,100000))
```


### Untangling

Untangling
----------
* Foot width as a function of sex and foot length.
* SAT and school expenditures
* [House prices](../Activities/UntanglingHousePrices.pdf)
