March 6, 2013 Class Notes
=============

```{r include=FALSE}
require(mosaic)
options(na.rm=TRUE)
```


### Review of Partial Derivatives

A derivative is a way of describing how to (quantitative) variables are related: it converts a change in one variable into a change in the other.

Partial derivatives are derivatives that hold all other variables constant.

Relate to the two-variable polynomial: $f(x,y) = a_0 + a_1 x + a_2 y + a_3 x y + ...$

With an interaction coefficient, the derivative with respect to $x$ depends on $y$.



### Discussion

Sometimes you want partial derivatives, sometimes you don't.

#### Kids feet data.

The question is whether girls' shoes are narrower than boys' because girls' feet are narrower.  Address this.

#### Wages

* Is there discrimination on the basis of sex?  Build a potentially complicated model and look at the partial change with respect to sex?
* What's the effect of education?  If you increase education by one year, you ought to decrease experience by one year.  


### Choice of Covariates

Consider the models `A ~ B+C` or `A ~ B*C`

When you include the covariate C,  and look at the coefficient on B, you are essentially looking at the partial of A with respect to B.

If you want to look at the total effect, you can either ...
* leave C out of the model
* evaluate the model for the proposed change in B together with the consequent change in C.

### Example: Looking at the Used Car Price Models

```{r}
cars = fetchData("used-hondas.csv")
mod1 = lm( Price ~ Mileage, data=cars)
mod2 = lm( Price ~ Mileage*Age, data=cars)
amod = lm( Age ~ Mileage-1, data=cars) # Why suppress the intercept
```

A 10000 mile increase in mileage is associated with a 0.57 year increase in age

Looking at the size of the effect:
```{r}
coef(mod1)
coef(mod2)
coef(amod)
f1 = makeFun(mod1)
f2 = makeFun(mod2)
```
Looking at the coefficients from the two models, the first says that the relationship of price to mileage is to decrease by about 10 cents/mile.  For mod 2, for a 4-year old car, the partial of price with respect to mileage is about -8 cents/mile.
```{r}
f1(Mileage=40000)-f1(Mileage=50000)
f2(Mileage=40000,Age=4)-f2(Mileage=50000,Age=4)
f2(Mileage=40000,Age=4)-f2(Mileage=50000,Age=4.57)
```


### Activity

Look at the Saratoga house price data.  Build and evaluate appropriate models to answer these questions:

(1) What's the effect of adding a fireplace?

* Should you hold the other variables constant?

(2) What's the effect of adding a bedroom?

* Should you hold living area constant?

(3) What's the effect of partitioning off the living room to create a bedroom?

* Should you hold living area constant?

### Activity

[Total-vs-partial In-class activity](https://dl.dropbox.com/u/5098197/ISM/total-partial.pdf)

### Longitudinal running data

Running data: Compare the cross-sectional to the longitudinal data to get at how tie changes versus age.  Question: hold individual constant or not.

```{r cache=TRUE}
f = fetchData("Cherry-Blossom-Long.csv")
nrow(f)
sample(f,size=5)
f = subset(f, nruns>5)
nrow(f)
```

Two models
```{r cache=TRUE}
mod1 = lm( net ~ age, data=f)
mod2 = lm( net ~ age + name.yob, data=f)
```
Model 1 suggests a decrease of about 0.3 minutes per year of age, but Model 2 is much larger, 0.83 minutes per year of age.

```{r}
coef(mod1)
head( coef(mod2))
```
Note how substantially the age dependence differs depending on whether you are looking longitudinally or cross-sectionally.

### Grades and the GPA

```{r}
grades = fetchData("grades.csv")
g2pt = fetchData("grade-to-number.csv")
courses = fetchData("courses.csv")
grades = merge(grades, g2pt) # Convert letter grade to number
grades = merge(grades,courses)
```

Compute the GPA in the ordinary way:
```{r}
options( na.rm=TRUE )
head( mean(gradepoint ~ sid, data=grades) )
```

... or by a model
```{r}
conventional = coef( lm( gradepoint ~ sid-1, data=grades))
head( conventional )
```

What can we hold constant?  Department, level, class enrollment?
```{r}
adjusted = coef( lm( gradepoint ~ sid - 1 + dept + level + enroll, data=grades))
head( adjusted )
```

How do they compare?
```{r}
xyplot( conventional ~ adjusted[1:443], pch=20 )
```

Or in terms of class rank:
```{r}
xyplot( rank(conventional) ~ rank(adjusted[1:443]), pch=20 )
```

Suppose the cut-off for class rank was to be $\geq 150$.  There are students who pass by the adjusted criteria but fail by the unadjusted.
```{r}
xyplot( rank(conventional) ~ rank(adjusted[1:443]), pch=20 )
plotFun( y >=150 ~ x&y, add=TRUE )
plotFun( x >=150 ~ x&y, add=TRUE )
```

The "takes easy courses" index: a positive number indicates taking easy courses.  
```{r}
densityplot( ~rank(conventional)- rank(adjusted[1:443]))
```


