Bias in promotion?
==================

Imagine that there are 10 women and 15 men in a promotion pool.  Here's such a pool in R:
```{r}
females = rep(1,10)
males = rep(0,15)
pool=c(males,females)
```

We're going to study how likely it is to get five females out of eight promotions in such a pool, under the hypothesis that sex is not a factor.  (Such a hypothesis, that nothing is going on, is called a "Null Hypothesis".)

The following, `pick()` is an R function that samples without replacement from the pool
```{r}
pick <- function(n) {
  sum(sample(pool,n))
}
```

The `pick()` function was written to be general, to work for any size promotion group.  Our promotion group had 8 people.  Let's simulate a random selection of 8 in a world where the null hypothesis is true.
```{r}
pick(8)
```

The above is a possible result, but we don't know if it is a common or an unlikely result in the null-hypothesis world.  Let's run the simulation many times and see what we get:
```{r}
n=100
Captain<-replicate(n,pick(8))
```

We could look at this in several ways, e.g. a table
```{r}
table(Captain)
``` 
or a graphic such as a histogram:
```{r}
hist(Captain,col="lightblue",
     ylim=c(0,.4*n),
     breaks=seq(-1.5,10.5,1))
```

Ultimately, we're interested in how often, in the null-hypothesis world, we would get a result of five or more females:
```{r}
table( Captain >= 5 )
```

**Note in passing**.  I think this expression is a little more obscure and verbose than needed:
```{r}
length(which(Captain>4))/n
```

Of course, it's a matter of taste.  The point of `which()` is that it extracts the indices of the entries which satisfy the criterion.  To me, it's more natural to use the condition itself, e.g.
```{r}
mean( Captain >= 5 )
```

## Doing it in a `mosaic` style

There's nothing wrong with the above, but I think it has several layers of code that are not strictly needed, e.g. using 0,1 for the sex, using rep(), and so on.

Here's one way to do it in `mosaic`

First, load the package.  (All the stuff in the .Rmd file is there just to avoid filling the html file with the messages from loading mosaic.)
```{r error=FALSE,warning=FALSE,message=FALSE}
require(mosaic)
```

* Create the pool.  In doing so, stick with the convention of data consisting of cases and variables, organized into a table or data frame.  Here, there is just one variable, sex.  But in general there might be several.
```{r}
pool = data.frame(sex=c("F","F","F","F","F","F","F","F","F","F","M","M","M","M","M","M","M","M","M","M","M","M","M","M","M"))
```

* Make sure that we got it right:
```{r}
tally(~sex, data=pool)
```

It just takes one time of having to type that whole expression above, before students appreciate why `rep()` is nice:
```{r}
pool = data.frame(sex=c( rep("F",10), rep("M",15)))
```
But always check!

```{r}
tally(~sex, data=pool)
```

* Now, do one simulation where we select 8 people from the pool.
```{r}
sample(pool, 8)
```

If we do this many times, we'll get somewhat different results from trial to trial:
```{r}
sample(pool, 8)
sample(pool, 8)
sample(pool, 8)
```

* Our interest is in the number of females who are promoted.  Let's do three trials:
```{r}
tally(~sex, data=sample(pool,8) )
tally(~sex, data=sample(pool,8) )
tally(~sex, data=sample(pool,8) )
```

* To get a better ideal of the variation, let's repeat this many times and collect the results of the trials
```{r}
trials = do(1000)*tally( ~ sex, data=sample(pool,8))
head(trials)
```

What did we get? Let's see how often each possibility arose for a number of promoted women:
```{r}
tally( ~F, data=trials)
```

It looks like having 5 women promoted out of this pool is fairly likely to happen under the null hypothesis. 
```{r}
tally( ~ F >= 5, data=trials)
```
Out of the 1000 trials, there were 5 or more women more than 10% of the time. So, there's no reason to reject the null hypothesis.


