Math 155: Introduction to Statistical Modeling
====================
``` {r echo=FALSE}
acroChapter <- function(chap,access="ISMf12",probnum=NULL,label=chap){
  if (!is.null(probnum)) probnum <- paste("#",probnum,sep="")
  
  paste("<a href='",
              paste("https://dl.dropbox.com/u/5098197/Exercises/HTML/",
                    "SM2-Chapter-",chap,"-Problems/SM2-Chapter-",chap,
                    "-Problems-A.html?access=",access,probnum,sep=""),
              "'>",label,"</a>",sep="")
}
# A short form for individual problems.
PN <- function(probs){
  res <- ""
  for (probnum in probs){
    chap <- substr(probnum,1,nchar(probnum)-2)
    prob <- substr(probnum,nchar(chap)+1,100)
    label <- paste(chap,".",prob,sep="")
    foo <- acroChapter(chap=chap, probnum=probnum, label=label )
    res <- paste(res,", ",foo,sep="")
  }
  return(substr(res,3,10000)) # But get rid of the trailing comma.
  
}
```

Links: AcroScore [accounts](https://moodle.macalester.edu/mod/assign/view.php?id=12216) and [settings](http://datavis.math.macalester.edu:8080/AcroScore08/) : Assignments by Chapter `r I(acroChapter(chap=1))`, `r I(acroChapter(chap=2))`, `r I(acroChapter(chap=3))`, `r I(acroChapter(chap=4))`, `r I(acroChapter(chap=5))`, `r I(acroChapter(chap=6))`, `r I(acroChapter(chap=7))`, `r I(acroChapter(chap=8))`, `r I(acroChapter(chap=9))`, `r I(acroChapter(chap=10))`, `r I(acroChapter(chap=11))`, `r I(acroChapter(chap=12))`, `r I(acroChapter(chap=13))`, `r I(acroChapter(chap=14))`, `r I(acroChapter(chap=15))`, `r I(acroChapter(chap=16))`, `r I(acroChapter(chap=17))`, `r I(acroChapter(chap=18))`


 


``` {r echo=FALSE}
# Setting up the calendar material
makeDayLinkName <- function(month,day,year=2012){
  paste("Day-", year, "-", month, "-", day, sep="")
}
dd <- function(month,day,year=2012) {
  paste("<a target='self' href='#",makeDayLinkName(month,day,year),"'>",day,"</a>",sep="")
}
classDayContents <- function(weekday="", month, day, year=2012) {
  # put in the header and a bookmark
  paste("### ", 
        paste('<a name="',makeDayLinkName(month,day,year),'">',weekday," ", day," ", month,"</a>",sep=""))
}
todaysNotes <- function(fname,marker="notes") {
  paste("<a href='../Notes/", fname, "'>",marker,"</a>",sep="")
}
```
**Class Schedule**

        | : | M | W | F | : | M | W | F | : | M | W | F | : | M | W | F | : | M | W | F
--------|---|---|---|---|---|---|---|---|---|---|---|---|---|---|---|---|---|---|---|---
*Sept* | | | `r I(dd("Sept",5))` | `r I(dd("Sept",7))` | |`r I(dd("Sept",10))` |`r I(dd("Sept",12))` |`r I(dd("Sept",14))` | | `r I(dd("Sept",17))` |`r I(dd("Sept",19))` |`r I(dd("Sept",21))` | | `r I(dd("Sept",24))` |`r I(dd("Sept",26))` |`r I(dd("Sept",28))` | | | | 
*Oct* |  |  `r I(dd("Oct",1))` |  `r I(dd("Oct",3))` | `r I(dd("Oct",5))` | | `r I(dd("Oct",8))` |`r I(dd("Oct",10))` |`r I(dd("Oct",12))` | | `r I(dd("Oct",15))` |`r I(dd("Oct",17))` |`r I(dd("Oct",19))` | | `r I(dd("Oct",22))` |`r I(dd("Oct",24))` | 26 | | `r I(dd("Oct",29))` |`r I(dd("Oct",31))` |
*Nov* | | | | `r I(dd("Nov",2))` | | `r I(dd("Nov",5))` |  `r I(dd("Nov",7))` | `r I(dd("Nov",9))` | |  `r I(dd("Nov",12))` | `r I(dd("Nov",14))` | `r I(dd("Nov",16))` | | `r I(dd("Nov",19))` | `r I(dd("Nov",21))` | 23 | |  `r I(dd("Nov",26))` | `r I(dd("Nov",28))` | `r I(dd("Nov",30))` 
*Dec* | | `r I(dd("Dec",3))` | `r I(dd("Dec",5))` | `r I(dd("Dec",7))` | | `r I(dd("Dec",10))` 

`r I(classDayContents("Wed", "Sept", 5 ))`
**Topic**: Introduction to the course. Randomness, modeling, and inference.
**Reading**: 
* *Statistical Modeling* textbook §1.1, 1.2, 1.3
* A [recent newspaper report](http://www.nytimes.com/2012/08/21/health/how-well-you-sleep-may-hinge-on-race.html?_r=1) that frames some of the technical issues covered by the course.

**Assignment**: 
* Do the reading! 
* AcroScore set up
* Pre-course survey

**On Your Own**:
* Watch this one-hour documentary: [The Joy of Stats](http://www.gapminder.org/videos/the-joy-of-stats/) with Hans Rossling

**In-class**: Starting with R. (Write a description of this activity)
Notes: `r I(todaysNotes("2012-09-05.html", "instr"))`

`r I(classDayContents("Fri", "Sept", 7 )) # ==============================`
**Topic**: Data: cases, variables, tables, and codebooks
**Reading**: Statistical Modeling §1.4, §2.1, 2.2, 2.3, 2.4
**Assignment**:  
* Exercises: `r I(PN(c("1RQ",101,102,104,105,110 )))`
* Quick survey to do **before** class: [link](https://spreadsheets.google.com/embeddedform?formkey=dEV1cW1rTWhpLUhITVFNVGpnQjNqT3c6MQ&theme=0AX42CRMsmRFbUy04ZWQwMDYwMS02YjZhLTQ2ZjMtYjcyNy0zYWNlMzlmYTAxNmY&embedded=true&ifq)

Notes: `r I(todaysNotes("2012-09-07.html", "instr"))`

`r I(classDayContents("Mon", "Sept", 10 ))`
**Topic**: Descriptive statistics
**Reading**: 
* SM §2.5, SM §3.1, 3.2, 3.3
* [XKCD cartoon](http://xkcd.com/539/)

**Assignment**: 
* Exercises `r I(PN(c(111,112,"2RQ",202,204,301,323,353)))`
* Elevator experiment, `r I(PN(c(336)))`

Notes: `r I(todaysNotes("2012-09-10.html", "instr"))`

`r I(classDayContents("Wed", "Sept", 12 ))`
**Topic**: Descriptive statistics: variance
**Reading**: SM §3.4, 3.5, [XKCD cartoon](http://xkcd.com/833/)
**Assignment**: `r I(PN(c(209,214,222,302,303,304,305,306,308,311,312,315,317,318,319 )))`

Notes: `r I(todaysNotes("2012-09-12.html", "instr"))`

`r I(classDayContents("Fri", "Sept", 14 ))`
**Topic**: Groupwise Models
**Reading**: SM Ch. 4
**Assignment**: `r I(PN(c(314.316,"3RQ",403,404,405,406,410)))`
**In Class**:
* Weekly Quiz

Notes: `r I(todaysNotes("2012-09-14.html", "instr"))`

`r I(classDayContents("Mon", "Sept", 17 ))`
**Topic**: Model values, residuals, and variance.  Sampling and resampling.
**Reading**: SM Ch. 4 (review).  SM §5.1, 5.2
**Assignment**: `r I(PN(c(354,"4RQ",407,408,501,502)))`

`r I(classDayContents("Wed", "Sept", 19 ))`
**Topic**: Confidence intervals
**Reading**: SM §5.3, 5.4, 5.5, 5.7
**Assignment**: `r I(PN(c("5RQ",503,512,517,520)))`
**In Class**: 
**On Your Own**: Study Guide to Friday's Quiz on R (link to be posted)

`r I(classDayContents("Fri", "Sept", 21 ))`
**Topic**: Language of Models
**Reading**: SM §6.1, 6.2, 6.3, 6.4
**Assignment**: 
* Required: `r I(PN(c(513,523, 530, 531,601,602,604,611,613)))`
* Optional, for the theoretically oriented.  `r I(PN(c(540)))`

**In Class**:
* Quiz on R

`r I(classDayContents("Mon", "Sept", 24 ))`
**Topic**: From Model Terms to Formulas and Coefficients
**Reading**: SM §6.5, 6.6, SM §7.1, 7.2
**Assignment**: 
* Required `r I(PN(c("6RQ",605,610, 612, 701, 702 )))`
* Extensions `r I(PN(c(620,621)))`, [XKCD cartoon](http://xkcd.com/418/).  What regression model for this?


`r I(classDayContents("Wed", "Sept", 26 ))`
**Topic**: Formulas and Coefficients
**Reading**: 
* SM §7.3, 7.4, 7.5, 7.6, 7.7, 7.8, 7.9
* [XKCD cartoon](http://xkcd.com/605/)

**Assignment**: 
* Exercises `r I(PN(c("7RQ", 704, 705,710,711,712, 713,715)))`
* Car project (group project): `r I(PN(c(723)))`

`r I(classDayContents("Fri", "Sept", 28 ))`
**Topic**: Fitting and Least Squares
**Reading**: SM §7.10, SM §8.1, 8.2 
**Assignment**: `r I(PN(c(714,716,721,722, 731, 801)))`
**In Class**: Weekly Quiz

`r I(classDayContents("Mon", "Oct", 1 ))`
**Topic**: Geometry of Least Squares Fitting
**Reading**: SM §8.3
**Assignment**: `r I(PN(c(802, 804, 805, 811)))`

`r I(classDayContents("Wed", "Oct", 3 ))`
**Topic**: Redundancy and Colinearity
**Reading**: SM §8.4, 8.5
**Assignment**:  `r I(PN(c("8RQ",805,812)))`
**In Class**: Redundancy activity `r I(PN(c(820)))`

`r I(classDayContents("Fri", "Oct", 5 ))`
**Topic**: Correlation and Partitioning, $R^2$
**Reading**: SM Ch. 9
**Assignment**: `r I(PN(c(901,902,904,905)))`
**In Class**: Weekly Quiz

`r I(classDayContents("Mon", "Oct", 8 ))`
**Topic**: Correlation and Partitioning (continued)
**Reading**: SM Ch. 9 (review)
**Assignment**: `r I(PN(c("9RQ",910,911,912,913,921)))`

`r I(classDayContents("Wed", "Oct", 10 ))`
**Topic**: Total and Partial Relationships
**Reading**: SM Ch. 10
**Assignment**: `r I(PN(c(922, 923,1001,1004,1010)))`
**In Class**: `r I(PN(c(1002)))`

`r I(classDayContents("Fri", "Oct", 12 ))`
**Topic**: Total and Partial Relationships (continued)
**Reading**: SM Ch. 10 (review)
**Assignment**: `r I(PN(c("10RQ",1005,1011,1012,1020)))`
**In Class**: Weekly Quiz

`r I(classDayContents("Mon", "Oct", 15 ))`
**Topic**: Confidence Intervals on Coefficients
**Reading**: SM Ch. 12
**Assignment**: `r I(PN(c(1021,1201,1202,1203,1205)))`

`r I(classDayContents("Wed", "Oct", 17 ))`
**Topic**: Review for Exam
**Reading**:
**Assignment**: Review Problems (to be posted)

`r I(classDayContents("Fri", "Oct", 19 ))`
**Mid-term Exam**

`r I(classDayContents("Mon", "Oct", 22 ))`
**Topic**: Confidence Intervals on Coefficients (continued)
**Reading**: SM Ch. 12
**Assignment**: `r I(PN(c("12RQ",1210,1212)))`

`r I(classDayContents("Wed", "Oct", 24 ))`
**Topic**: Probability Models
**Reading**: SM Ch. 11
**Assignment**: `r I(PN(c("11RQ",1101,1102,1104,1105,1110,1121,1122)))`

`r I(classDayContents("Fri", "Oct", 26 ))`
*FALL BREAK*

`r I(classDayContents("Mon", "Oct", 29 ))`
**Topic**: Hypothesis Testing
**Reading**: SM Ch. 13
**Assignment**: `r I(PN(c(1111,1130,1135,1137,1301)))`

`r I(classDayContents("Wed", "Oct", 31 ))`
**Topic**: Hypothesis Testing
**Reading**: SM Ch. 13 (review)
**Assignment**: `r I(PN(c("13RQ",1302)))`
**In Class**: `r I(PN(c(1315)))`

`r I(classDayContents("Fri", "Nov", 2 ))`
**Topic**: Testing Whole Models with $latex R^2$ and F
**Reading**: SM §14.1, 14.2, 14.3, 14.5
**Assignment**: `r I(PN(c(1401,1402)))`
**In Class**: 
* Weekly Quiz
* `r I(PN(c(1422)))`

`r I(classDayContents("Mon", "Nov", 5 ))`
**Topic**: Interpreting the p-value
**Reading**: 
* SM §14.4
* [XKCD cartoon](http://xkcd.com/882/) and [another](http://xkcd.com/628/)

**Assignment**: `r I(PN(c("14RQ",1404,1405,1411,1412,1424)))`
**In Class**: `r I(PN(c(1421)))`


`r I(classDayContents("Wed", "Nov", 7 ))`
**Topic**: Hypothesis Testing on Parts of Models
**Reading**: SM Ch. 15
**Assignment**: `r I(PN(c()))`

`r I(classDayContents("Fri", "Nov", 9 ))`
**Topic**: Hypothesis Testing on Parts of Models (continued)
**Reading**: SM Ch. 15 (review)
**Assignment**: `r I(PN(c()))`
**In Class**: Weekly Quiz

`r I(classDayContents("Mon", "Nov", 12 ))`
**Topic**: Choosing model terms
**Reading**: [XKCD cartoon](http://xkcd.com/675/)
**Assignment**: `r I(PN(c()))`

`r I(classDayContents("Wed", "Nov", 14 ))`
**Topic**: Non parametrics
**Reading**:
**Assignment**: `r I(PN(c()))`

`r I(classDayContents("Fri", "Nov", 16 ))`
**Topic**: Logistic regression
**Reading**: SM Ch. 16
**Assignment**: `r I(PN(c()))`
**In Class**: Weekly Quiz

`r I(classDayContents("Mon", "Nov", 19 ))`
**Topic**: Logistic regression
**Reading**: SM Ch. 16
**Assignment**: `r I(PN(c()))`

`r I(classDayContents("Wed", "Nov", 21 ))`
**Topic**: Logistic regression
**Reading**: SM Ch. 16
**Assignment**: `r I(PN(c()))`

`r I(classDayContents("Fri", "Nov", 23 ))`
*THANKSGIVING BREAK*

`r I(classDayContents("Mon", "Nov", 26 ))`
**Topic**: Causation
**Reading**: 
* SM Ch. 17
* [XKCD cartoon](http://xkcd.com/925/)

**Assignment**:

`r I(classDayContents("Wed", "Nov", 28 ))`
**Topic**: Causation
**Reading**: 
* SM Ch. 17
* [XKCD cartoon](http://xkcd.com/552/)

**Assignment**:

`r I(classDayContents("Fri", "Nov", 30 ))`
**Topic**: Causation
**Reading**: SM Ch. 17 and [XKCD cartoon](http://xkcd.com/687/)
**Assignment**: `r I(PN(c("17RQ")))`
**In Class**: Weekly Quiz

`r I(classDayContents("Mon", "Dec", 3 ))`
**Topic**: Experiment and Design
**Reading**: SM Ch. 18
**Assignment**:

`r I(classDayContents("Wed", "Dec", 5 ))`
**Topic**: Experiment and Design
**Reading**: SM Ch. 18
**Assignment**:

`r I(classDayContents("Fri", "Dec", 7 ))`
**Topic**: Experiment and Design
**Reading**: SM Ch. 18
**Assignment**: `r I(PN(c("18RQ")))`

`r I(classDayContents("Mon", "Dec", 10 ))`
**Topic**: Course Review and Wrap-Up
**Reading**:
**Assignment**:
