Stats 155 Class Notes 2012-11-05
================================

```{r name="setup", child="notes-setup.Rmd"}
# boilerplate
```

### Hypothesis testing framework review

The Ivory Billed woodpecker from a [hypothesis testing perspective](bayes-approach.pdf) second page.

### Reminder about confidence intervals

Confidence intervals are computed on Planet Samp.

1. What fraction of the earth is covered by water?  Toss the globe around.
2. How much of the US is within one mile of a road?  
    * Use google maps with a location specified like this (100.342 W, 35.873 N)  and look for the green arrow rather than the "push-pin" (which gives the nearest address, which is typically near a road).
    *Pick a random longitude between 65 and 125 degrees W, and a latitude between 25 and 50 degrees N.  If you get water, pick a new point.  (Here's a [map](USmap.png) that can be used to justify this range of locations.)
    !<USmap.png>
    * Tally up how many yesses and no's.
    * Resample these to get a standard error.  Also do the regression report.

### Shuffling 

[Permutation test by hand](http://dl.dropbox.com/u/5098197/ISM/Activities/permutation-test.pdf)

### Power

How did I know that I could reject the Null in the shuffling problem?  I did a little simulation.
```{r}
mysim <- function(n=15) {
   days = resample(1:31,size=n)
   nums = ceiling( runif(n,min=0,max=days))
   mod = lm(nums~days)
   list(r2=r.squared(mod), p=summary(mod)$coef[2,4])
}
s15 = do(1000)*mysim(15) # typical R^2 is about 0.4
mean(~r2, data=s15)
tally( ~ p<0.05, data=s15,format='proportion')
```

Do `mHypTest(TRUE)` setting the "effect size" to about 0.4

### Interpreting p-values

### Activity

Print out list of 20 p-values. Ask who has got the first one < 0.05? Who has got any of them? What's the smallest p-value in the class?

Simulation activity:
```{r}
fetchData("simulate.r")
d = run.sim( bogus.groups, ngroups=5, 10)
bwplot(val~group, data=d)
dd = subset( d, group %in% c("A", "E") ) # or whatever are the furthest-apart groups
```

Carry out a hypothesis test on a subset that is just the best and worst groups. How often do people get a p-value less than 0.05?

Publication bias.  Filing cabinet problem.

### Meta-studies and funnel plots 

* The standard error and the funnel plot: looking at lots of studies and comparing their odds ratios and standard errors: Start at time 9:30 in [this TED video](http://www.ted.com/talks/ben_goldacre_battling_bad_science.html) 
* [Wikipedia](http://en.wikipedia.org/wiki/Funnel_plot)
