Stats 155 Class Notes 2012-11-30
================================

```{r name="setup", child="notes-setup.Rmd"}
# boilerplate
```

**In the News**
* [No one shot or stabbed in NY on Monday](http://www.npr.org/blogs/thetwo-way/2012/11/28/166083278/this-just-in-no-one-was-shot-or-stabbed-monday-in-new-york-city?sc=ipad&f=1001)  We know the annual murder rate: about 400/365 per day.  So there are many days when there will be no murders.  What must the "stabbing" rate be so that it would be a surprise that there were no murders or stabbings?  A poisson model.

#### Confounders as a diagram

<pre>
Variable   --->  Outcome
   ^                 ^
   |                 |
   ---> Confounder <--
</pre>

#### SAT versus per-student spending

Draw the causal diagram

<pre>
Spending  ---------->      SAT    <-------|
      |                                   |
Focus on Educ. ---> fraction taking SAT  -|
</pre>

#### Campaign Spending and the back-door network

Research in political science shows that higher spending in campaigns is related to a lower vote for the incumbent.  Yet it's common sense that higher spending improves things for the candidate; that's why they do it.

<pre>
  Polls <-----    Popularity ---> vote outcome
    |                                 ^
    v                                 |
  Spending ---------------------------
</pre>

### How to block a back-door?

We've done it by including the covariate in the model.  But this is too crude an answer.

### Example: Election Spending

```{r}
fetchData("simulate.r")
campaign.spending
equations(campaign.spending)
```
You can see from the equations how spending is related to vote: it increases it.  Let's look at what a statistical model has to say
```{r}
d = run.sim(campaign.spending, 435 ) # number of congressment
summary( lm( vote ~ spending, data=d))
```
The problem: a back door pathway from spending to vote via popularity.  Block it by including a node on the pathway as a covariate.

```{r}
summary( lm( vote ~ spending + polls, data=d))
```
Work session on Barry Bonds at Bat
----------------

Work on the Logistic Regression model of Bonds's hitting.
