Stats 155 Class Notes 2012-12-03
================================

```{r name="setup", child="notes-setup.Rmd"}
# boilerplate
```
$latex \newcommand{\ensuremath}[1]{#1}
\newcommand{\pathway}[1]{\hbox{#1}}
\newcommand{\causes}{\ensuremath{\Rightarrow}}
\newcommand{\causedBy}{\ensuremath{\Leftarrow}}
\newcommand{\correlatedWith}{\ensuremath{\nLeftrightarrow}}$


### Software for Simulation

You should do this even if you already have done so.  There were some corrections made to the software.

```{r}
fetchData("simulate.r")
```

### Chains in Causation

Following the convention used throughout the book
* $A$ is a response variable
* $B$ is an explanatory variable
* $C$ is a covariate.  We'll also use $D$, $E$, etc. as covariates and $U$ or $?$ as unobserved variables.

#### Three linear pathways:
* **Causal mediator**: $latex A \causedBy C \causedBy  B$.  Simulation: `cnet.mediator`
* **Common cause**: $latex A \causedBy C \causes B$. Simulation: `cnet.common.cause`
* **Witness**: $C$ is caused by both $A$ and $B$: $latex A \causes C \causedBy B$. Simulation: `cnet.witness`

#### A circular pathway
* A **recurrent network** (or, closed loop of causation): : $latex A \causes B \causes C \causes A$ or the variations.

We won't deal with these.  They raise questions about "when".

#### Correlation

The notation $latex A \correlatedWith B$ means that there is a non-causal connection between $latex A$ and $latex B$.  This must be because there is some unobserved variable $U$ producing the correlation, such as 

1. $latex A \causedBy U \causes B$
2. $latex A \causes U \causedBy B$

Note that if either of the arrows went the other way, we'd simply have $U$ being a causal mediator.

We'll see later that (1) is not a possibility if $U$ is unobserved and unconnected to anything else.

### The Rules for Observing or Blocking a Pathway

Remember, we want to study the **causal** relationship between $latex A$ and $latex B$ and are trying to decide whether to include the covariate $latex C$.

* Causal mediator: Including $latex C$ **blocks** the pathway, which is otherwise open.
* Common cause: Including $latex C$ **blocks** the pathway, which is otherwise open.
* Witness: Including $latex C$ **opens** the pathway, which is otherwise blocked.

### Confirming these rules

Run a simulation of $latex A$ against $latex B$.  When the coefficient in a model is zero, the pathway between $latex B$ and $latex A$ has been blocked. 

#### Mediator

```{r}
equations(cnet.mediator)
s = run.sim(cnet.mediator, 1000 )
coef( summary( lm( A ~ B, data=s )))
```
The above model indicates that the pathway is open.

Including the covariate $latex C$ blocks the pathway between $latex B$ and $latex A$.
```{r}
coef( summary( lm( A ~ B + C, data=s )))
```

### Common Cause

```{r}
equations(cnet.common.cause)
s = run.sim(cnet.common.cause, 1000 )
coef( summary( lm( A ~ B, data=s )))
```
The above model indicates that the pathway is open.

Including the covariate $latex C$ blocks the pathway between $latex B$ and $latex A$.
```{r}
coef( summary( lm( A ~ B + C, data=s )))
```

### Witness

```{r}
equations(cnet.witness)
s = run.sim(cnet.witness, 1000 )
coef( summary( lm( A ~ B, data=s )))
```
The above model indicates that the pathway is blocked.

Including the covariate $latex C$ opens the pathway between $latex B$ and $latex A$.
```{r}
coef( summary( lm( A ~ B + C, data=s )))
```

### The Jock Network 

Imagine that we wanted to study the relationship between athletic ability and SAT scores.  The conventional wisdom is that athletes tend to have lower test scores.  The `jock` network implements one hypothesis.

```{r}
s = run.sim(jock, 1000)
head(s)
```

In practice, SAT and athletic ability data on an individual might be available via a college, so let's pull out just the cases that are in college:
```{r}
college <- subset(s, College=="Yes")
```

Look at the relationship between IQ and Athletic:
```{r}
coef(summary(lm( IQ ~ Athletic, data=college)))
```

There is a relationship!

Now look at the entire set of data:
```{r}
coef(summary(lm(IQ~Athletic, data=s)))
```
The relationship has disappeared.  Why?

ACTIVITY: Look at the structure and equations of the `jock` network and explain.

### In-Class Activity

Here is the complete set of simple causal pathways among three variables.

* In each case, write down whether the pathway is a mediator, a common cause, or a witness.
* We've only studied the pathways with C in the middle.  For the others, figure out whether including C as a covariate opens or blocks the pathway between A and B.
* Come up with a setting which assigns a meaning to A, B, and C.  Example: B is taking nitroglycerin, C is dilation of the arteries, A is getting a headache.  

The possibilities for who can be in the middle: $latex A$ or $latex B$ or $latex C$

#### $C$ in the middle

* $latex B \causedBy C \causes A$
* $latex B \causedBy C \causedBy A$
* $latex B \causes C \causedBy A$
* $latex B \causes C \causes A$

#### $A$ in the middle

* $latex B \causedBy A \causes C$
* $latex B \causedBy A \causedBy C$
* $latex B \causes A \causedBy C$
* $latex B \causes A \causes C$

#### $B$ in the middle

* $latex A \causedBy B \causes C$
* $latex A \causedBy B \causedBy C$
* $latex A \causes B \causedBy C$
* $latex A \causes B \causes C$


Barry Bonds at Bat
===================

Analyzing the Survey Data
