Stats 155 Class Notes 2012-09-21
========================================================

```{r name="setup", child="notes-setup.Rmd"}
# boilerplate
```

## Review
Groupwise models
* Easy enough to fit groupwise models.  The groupwise mean is a nice estimate for the parameter that eliminates the bias in the residuals and minimizes the variance.
* We can use more than one categorical variable as a grouping variable.  When we do this, we get results for all the "crosses".

**BUT ...**
* What if we want to use a quantitative variable as an explanatory variable?
* What if we have more than one categorical variable?
* What if we have a mixture of quantitative and categorical variables?

Why would we want to use more than one variable?  Because things are complicated.

Example: Do wages depend on sex in the CPS85 data?
```{r}
mod = mm( wage ~ sex, data=CPS85 )
coef(mod) # men make about $2.20 per hour more
summary(mod) # the confidence intervals don't overlap
```

We could find the resampling distribution on the difference between men's and women's wages using the techniques we talked about in the last class.  Just for convenience, we can use the `diff()` function to find the difference between the two groups:
```{r cache=TRUE, out.width="2in"}
diff( coef(mod) )  # The point estimate
# Do some resampling to get the resampling distribution
trials = do(100)*diff( 
  coef( mm(wage ~ sex, data=resample(CPS85))))
qdata(c(0.025, .975), M, data=trials)
sd( trials) # Standard error
densityplot( ~ M, data=trials )
```
So there seems to be a difference between the men's and women's average wages. 
 
#### Possible other factors

Ways in which men and women might differ:
* education
* age
* type of job

How could we look at the issue of differences between the sexes in age, taking into account these other possibilities?

Sector of the economy is easy:
```{r}
summary(mm(wage~sex+sector, data=CPS85))
```
Compare the corresponding sector for the different sexes.  The confidence intervals often overlap.

How to add in additional variables, e.g. age and education?


The basic models
-----------------

Assume we have a response variable D, a quantitative explanatory variable, A, and a categorical variable G.
Here are the basic types of models, illustrated on the swimming data. 

More complicated models mainly add additional variables but follow the same overall pattern.

```{r}
swim = fetchData("swim100m.csv")
```

* `D ~ 1`  

All cases are the same.  There is one parameter, the grand mean.
```{r fig.height=4}
mod = lm( time ~ 1, data=swim)
xyplot( time + fitted(mod) ~ year, data=swim)
```

* `D ~ 1 + A`
The cases vary with the quantitative variable.  There are two parameters: slope and intercept.
```{r fig.height=4}
mod = lm( time ~ 1 + year, data=swim)
xyplot( time + fitted(mod) ~ year, data=swim)
```

* `D ~ 1 + G`
The groupwise mean model.  But note that the parameters are written differently, as an "intercept" and a "difference".  QUESTION: Which difference is it?
```{r fig.height=4}
mod = lm( time ~ 1 + sex, data=swim)
xyplot( time + fitted(mod) ~ year, data=swim)
```

* `D ~ 1 + A + G`
Both variables play a role.  The parameters are "intercept," "slope," and "difference between groups."

```{r fig.height=4}
mod = lm( time ~ 1 + year + sex, data=swim)
xyplot( time + fitted(mod) ~ year, data=swim)
```
Note that the slope is the same for the two groups.

To see a different slope for the two groups, we have to ask the model to permit such a thing:

* `D ~ 1 + A + G + A:G`
```{r fig.height=4}
mod = lm( time ~ 1 + year + sex + year:sex, data=swim)
xyplot( time + fitted(mod) ~ year, data=swim)
```

### Variables and Model Terms
How a variable is different from a model term.
Model terms that are not variables: intercept, interaction terms.

### Activity
Visual choice of model terms

Instructions are [given here](http://dl.dropbox.com/u/5098197/Exercises/HTML/ISM-Selecting-Model-Terms/ISM-Selecting-Model-Terms-A.html?access=ISM3)  (These have to be reformatted as a handout and as an instructor's guide.)

The only substantive differences is in the way the software is loaded:
```{r}
fetchData("mLM.R")
```

The program itself is called `mLM()` (not `lm.select.terms()`)





### Weekly quiz


