In-Class Notes: 2012-09-19
========================================================

```{r name="setup", child="notes-setup.Rmd"}
# boilerplate
```

### Review from last class

We saw that the choice of parameters in a model could be automated: fit the model, thereby choosing the parameters that eliminate bias in the residuals and minimize the variance.

Of course, the chosen parameters depend on the data to which the model is fitted.  Insofar as this data is a random sample from a population, the parameters themselves are random.  The question for use today is, "How random?"

### The sampling process

Sample 100 runners from the runners data set and find the mean running time.  Because there is some missing data, you'll need to do something like this:
```{r}
options(na.rm=TRUE)
```


Across the class, see what the range of values is.

Do it several times on the instructor's computer.

Do it many times with do and plot out the sampling distribution.


### Resampling

You don't typically have the population at hand.  (If you did ... )

So we simulate the sampling distribution by constructing a simulated population that consists of many copies of the original.

### The Standard Error

### From the Sampling Distribution to a Confidence Interval

### Formats for the Confidence Interval

### Interpreting the Confidence Interval

#### Differences between groups

Simple visual method: do the confidence intervals overlap?  Later on we'll see a more sophisticated method called hypothesis testing.  

#### Coverage
Have everyone construct a groupwise model of the wage data against some grouping variable and pick some group to examine.
* First, do it to the entire data set, as if it were a population.
* Second, do it to a sample of 100 cases.
* Construct a 50% confidence interval.

Is the population value within your confidence interval?

### Significant digits and the confidence interval