2012-09-12 Class Notes
--------------------------

```{r error=FALSE,warning=FALSE,message=FALSE,echo=FALSE,results="hide"}
require(mosaic, quietly=TRUE)
```


### Measurement and measurement bias
* Instrumental or procedural error. Measure heights of people with a ruler, but with the ruler not level.
* Subject or response bias: "How many of you are intravenous drug users?"
* Global warming. Measurement gauges have been encroached on by urbanization.


### Sampling and sampling bias
* Self-selection bias for surveys
* Problems with convenience samples: e.g., interview students at library on Friday night about their study habits
* Selection and survival bias: we don't see the kids who drop out of school, tracking a random sample of Alzheimer's patients to see how long they survive --- you're more likely to sample those who survive a long time.

### Random sampling
* Sampling frame and formal methods

#### In-Class Activity

* Sampling Bias Activity `s = select.books()` which needs to be proceeded by
```{r}
fetchData("simulate.r")
```

[Instructor's write up](../Activities/SelectingLibraryBooks.html)

## Graphics about distributions 

(from [Monday](2012-09-10.html))


## Numerical summaries of distributions (from [Monday](2012-09-10.html))

mean, variance, and standard deviation

other measures (e.g., IQR, median, 95% coverage)

[In-class activity]((https://dl.dropbox.com/u/5098197/ISM/Numerical-Summaries.pdf) on the properties of the various measures.

### What do your fellow students know about statistics
```{r}
ks = fetchData("/Users/kaplan/Dropbox/Stat155Fall2012/knowledge-survey-2012-09-11.csv")
for (item in levels(ks$ProblemItem)) {
  print( item )
  print(tally(~as.character(AnswerContents), data=subset(ks, ProblemItem==item)))
  print("")
}
```


