Stats 155 Class Notes 2012-10-10
================================

```{r name="setup", child="notes-setup.Rmd"}
# boilerplate
```





Review
=======
$R^2$ between zero and 1.  
* Use delta with nested models to decide if a model is better than junk.
* SAT versus GPA, $R^2 \approx 0.16$.  Based on thousands of student records.
* Little $r$ and the meaning of $r=0.40$

Question about grades: How would you decide if the SAT is deficient or whether it is as good as it can possibly be and grades are deficient?

Survey Project
===============

1. Hand out [project description](http://dl.dropbox.com/u/5098197/Exercises/HTML/ISM-ISM-Spring-2011-Term-Project/ISM-ISM-Spring-2011-Term-Project-A.html?access=ISM3) They are going to be creating a Google Form to collect the data.
2. Ask for suggestions about project areas
3. Have students form groups around common interests. 


Partial versus total
===============

Suppose I have a used car.  I'm going away for a year and thinking of selling it.  On the other hand, it would be nice to have a car available when I get back.  How much will it cost me to delay selling the car for a year?

Consider several models of used car prices fitted to data on used Hondas:
```{r}
cars = fetchData("used-hondas.csv")
mod1 = lm( Price ~ Age, data=cars )
mod2 = lm( Price ~ Mileage, data=cars )
mod3 = lm( Price ~ Mileage + Age, data=cars )
mod4 = lm( Price ~ Mileage * Age, data=cars )
```

Each of the first three is nested in the 4th.  You can play around with the various models this way:
```{r eval=FALSE}
fetchData("mLM.R")
mLM( Price ~ Age*Mileage, data=cars )
```
Include and exclude terms to try to answer this question:
> Which is the right model to use to inform my car-selling decision?

Tempting to use model 1, since Age is the only variable that I'm interested in.
```{r}
xyplot( fitted(mod1)+Price ~ Age, data=cars )
```

It's a bit hard to see the model.  Let's try another way of plotting it.
```{r}
f1 = makeFun(mod1)
plotPoints(Price ~ Age, data=cars)
plotFun( f1(Age)~ Age, add=TRUE)
```
How much the price goes down with a year depends on how old the car is, but you can get the rate from the derivative of the function.  Let's evaluate that derivative for an 8-year old car with 50,000 miles:
```{r}
f1 = makeFun(mod1)
df1 = D(f1(Age)~Age)
df1(Age=8)
```
Or, since I'm really thinking about a 1-year difference:
```{r}
f1( Age=c(9,8), Mileage=50000 )
```
Take the difference.
QUESTION: How come I get the same answer for the finite-difference and the derivative?

But let's consider mod4
```{r}
f4 = makeFun(mod4)
plotFun( f4(Age=a,Mileage=m)~a&m, 
         a.lim=c(0,10), m.lim=c(0,100000),
         levels=1000*(1:20),npts=200)
plotPoints( Mileage ~ Age, data=cars,add=TRUE,
            pch=20, col="red")
```

Examine the change as age goes up by one year.  Should I hold mileage constant or should I let the mileage change with age in the typical way?  

Here's the same question another way: Do I want to compare cars with different mileages and different ages, or do I want to compare cares with different ages and the same mileage.

Calculating the partial derivative or partial change:
* Take a finite-difference approach 
```{r}
f4( Age=c(8,9),Mileage=50000)
```

* Use partial derivatives
```{r}
df4da = D( f4(Age=Age,Mileage=Mileage)~Age)
df4da(Age=8, Mileage=50000 )
```

* Do it analytically, using the coefficients to construct the model polynomial.


### Review of Partial Derivatives

Relate to the two-variable polynomial: $f(x,y) = a_0 + a_1 x + a_2 y + a3 x y + ...$

Relate to the three-variable polynomial.

* Calculation from model function by evaluating at nearby points

### Real-world examples

* Insurance example: Should we pay people not to be on the College's health plan?
* NGOs and development aid: Speaker Christie Hansen talked about how donors for health funding in Kenya ended up funding other activities as the government moved money away from health toward agriculture, etc.
* Merit aid. Will it reduce net tuition or increase it. Depends on whether we substantially change the yield.

### Activity

[Total-vs-partial In-class activity](https://dl.dropbox.com/u/5098197/ISM/total-partial.pdf)

Car data: work with the two-dimensional graphs