Stats 155 Class Notes 2012-11-02
================================

```{r name="setup", child="notes-setup.Rmd"}
# boilerplate
```

### Stock Prices

Let's start by taking a trip to planet Sample.  We'll look at stock prices.  

For now good reason, there's a two-step process for reading in the stock data:
```{r}
fetchData("getDJIAdata.R")
djia = getDJIAdata()
```

Economists often look at "log prices," since changes are often proportional (think how inflation rates are given).  

If we invest for 10 years (about 2500 days), what kind of return will we get?

```{r}
new = subset(djia, Date > as.Date("1990-01-01"))
daily = diff( log(with(data=new, Close)))
```

Now a sample of 2500 days
```{r}
sim = exp(sum(resample(daily,size=2500)))
```

Or, many:
```{r}
s = do(1000)*exp(sum(resample(daily,size=2500)))
densityplot(~result, data=s)
```

Looking at the distribution of yearly returns:
```{r}
yearly = do(1000)*exp(sum(resample(daily,size=250)))
densityplot(~result, data=yearly)
```
What's the chance that you'll loose money in any given year?

This analysis assumes that the daily returns are independent.  Are they.

```{r}
fetchData("takeLag.R")
new = transform(new, change=(Close-Open)/Close)
oneday = takeLag(new)
```

How would you decide if there is any relationship between one day and the next?

### Drunkard's walk and hypothesis test

### Ivory Bill done as a hypothesis test



### R^2 as a test statistic

Activity: fit the kids' feet data with different numbers of model terms: `rand(p)` 

Find an average R^2 for each p and draw the graph of R^2 versus p.

### From R^2 to F

Construction of F

Comparing the distribution of F to the distribution of R^2.  F is much more stable, easier to interpret.  We look for F near 1 if the null is true.

### Power and Significance activity.

