2012-09-12 Class Notes
--------------------------
```{r error=FALSE,warning=FALSE,message=FALSE,echo=FALSE,results="hide"}
require(mosaic, quietly=TRUE)
```

### What do your fellow students know about statistics
```{r}
ks = fetchData("/Users/kaplan/Dropbox/Stat155Fall2012/knowledge-survey-2012-09-11.csv")
for (item in levels(ks$ProblemItem)) {
  print( item )
  print(tally(~as.character(AnswerContents), data=subset(ks, ProblemItem==item)))
  print("")
}
```

### Measurement and measurement bias
* Instrumental or procedural error. Measure heights of people with a ruler, but with the ruler not level.
* Subject or response bias: "How many of you are intravenous drug users?"
* Global warming. Measurement gauges have been encroached on by urbanization.


### Sampling and sampling bias
* Self-selection bias for surveys
* Problems with convenience samples: e.g., interview students at library on Friday night about their study habits
* Selection and survival bias: we don't see the kids who drop out of school, tracking a random sample of Alzheimer's patients to see how long they survive --- you're more likely to sample those who survive a long time.

### Random sampling
* Sampling frame and formal methods

#### In-Class Activity

* Sampling Bias Activity `s = select.books()` which needs to be proceeded by
```{r}
fetchData("simulate.r")
```

[Instructor's write up](../Activities/SelectingLibraryBooks.html)

## Graphics about distributions (from [Monday](2012-09-10.html))

## Numerical summaries of distributions (from [Monday](2012-09-10.html))

mean, variance, and standard deviation

other measures (e.g., IQR, median, 95% coverage)

[In-class activity]((https://dl.dropbox.com/u/5098197/ISM/Numerical-Summaries.pdf) on the properties of the various measures.



### The Knowledge Survey collected half-way through
Wednesday night, with 64 responses
<pre>
Summary Report for problem Prob-knowledge-survey

> summarize.problems( consolidate.files( filenames ), show.vals=TRUE)

Prob-knowledge-survey.algebra AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12          7
  AP\n   course  1
  College       35
  Never         21

Prob-knowledge-survey.ANCOVA AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores      
        0
  None 48
  Some 16

Prob-knowledge-survey.ANOVA AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  1
  None     32
  Some     31

Prob-knowledge-survey.bayes AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  4
  None     47
  Some     12

Prob-knowledge-survey.boot AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores      
        0
  None 61
  Some  2

Prob-knowledge-survey.bwplot AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12          9
  4-6           15
  6-9           23
  AP\n   course  4
  College        2
  Never         11

Prob-knowledge-survey.calc AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         20
  4-6            3
  6-9           31
  AP\n   course  4
  College        1
  Never          4

Prob-knowledge-survey.coll AJAX                     -
   62 answers from   62 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  1
  None     46
  Some     15

Prob-knowledge-survey.cond AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  1
  None     50
  Some     12

Prob-knowledge-survey.confI AJAX                     -
   62 answers from   62 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  7
  None     35
  Some     20

Prob-knowledge-survey.cprob AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         19
  4-6            3
  6-9            9
  AP\n   course  4
  College        1
  Never         28

Prob-knowledge-survey.cumu AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  3
  None     44
  Some     16

Prob-knowledge-survey.DF  AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  3
  None     33
  Some     27

Prob-knowledge-survey.dummy AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  1
  None     46
  Some     16

Prob-knowledge-survey.excel AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         20
  4-6            1
  6-9           19
  AP\n   course  2
  College       12
  Never         10

Prob-knowledge-survey.exp AJAX                     -
   62 answers from   62 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 20
  None     12
  Some     30

Prob-knowledge-survey.Ftest AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores      
        0
  None 60
  Some  3

Prob-knowledge-survey.hist AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores               
                 0
  1-3            3
  10-12         12
  4-6           15
  6-9           20
  AP\n   course  3
  College        1
  Never         10

Prob-knowledge-survey.Hnull AJAX                     -
   62 answers from   62 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 22
  None     20
  Some     20

Prob-knowledge-survey.intrct AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  1
  None     48
  Some     14

Prob-knowledge-survey.LA  AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         14
  4-6            5
  6-9           14
  AP\n   course  1
  College       11
  Never         19

Prob-knowledge-survey.lincomb AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 11
  None     35
  Some     17

Prob-knowledge-survey.logist AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores      
        0
  None 50
  Some 13

Prob-knowledge-survey.lreg AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 18
  None     20
  Some     26

Prob-knowledge-survey.LSQ AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  9
  None     35
  Some     19

Prob-knowledge-survey.main AJAX                     -
   62 answers from   62 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  1
  None     48
  Some     13

Prob-knowledge-survey.mmm AJAX                     -
   58 answers from   58 readers. Mean score 0.00
Answers vs Scores         
           0
  1-3      7
  10-12    1
  4-6     33
  6-9     14
  College  1
  Never    2

Prob-knowledge-survey.mTerm AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  1
  None     50
  Some     12

Prob-knowledge-survey.normal AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         29
  6-9           14
  AP\n   course  6
  College        7
  Never          8

Prob-knowledge-survey.ortho AJAX                     -
   62 answers from   62 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  9
  None     40
  Some     13

Prob-knowledge-survey.out AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 45
  None      5
  Some     14

Prob-knowledge-survey.perc AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 39
  None      1
  Some     24

Prob-knowledge-survey.prob AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores               
                 0
  1-3            5
  10-12         26
  4-6           13
  6-9           17
  AP\n   course  1
  College        1
  Never          1

Prob-knowledge-survey.program AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores               
                 0
  6-9            1
  AP\n   course  1
  College       24
  Never         38

Prob-knowledge-survey.pVal AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 12
  None     26
  Some     26

Prob-knowledge-survey.rand AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 15
  None     24
  Some     24

Prob-knowledge-survey.rank AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores      
        0
  None 57
  Some  6

Prob-knowledge-survey.rcoef AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 11
  None     17
  Some     36

Prob-knowledge-survey.regres AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         21
  6-9            6
  AP\n   course  7
  College       13
  Never         17

Prob-knowledge-survey.resid1 AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  9
  None     24
  Some     30

Prob-knowledge-survey.sampD AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  5
  None     25
  Some     33

Prob-knowledge-survey.scatter AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         11
  4-6           18
  6-9           26
  AP\n   course  3
  College        3
  Never          3

Prob-knowledge-survey.sd  AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores          
            0
  10-12     2
  6-9       1
  College   1
  Complete 30
  None      6
  Some     24

Prob-knowledge-survey.sig AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  8
  None     36
  Some     19

Prob-knowledge-survey.skew AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  7
  None     34
  Some     22

Prob-knowledge-survey.sResid AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  6
  None     39
  Some     18

Prob-knowledge-survey.subsp AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  8
  None     31
  Some     24

Prob-knowledge-survey.t   AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores               
                 0
  1-3            1
  10-12         11
  6-9            1
  AP\n   course  9
  College       13
  Never         29

Prob-knowledge-survey.tTest AJAX                     -
   64 answers from   64 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  4
  None     32
  Some     28

Prob-knowledge-survey.TypeI AJAX                     -
   63 answers from   63 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  2
  None     44
  Some     17

<pre>

Monday night: with 48 responses
<pre>
> summarize.problems( consolidate.files( filenames ), show.vals=TRUE)

Prob-knowledge-survey.algebra AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12          4
  AP\n   course  2
  College       26
  Never         16

Prob-knowledge-survey.ANCOVA AJAX                     -
   46 answers from   46 readers. Mean score 0.00
Answers vs Scores      
        0
  None 35
  Some 11

Prob-knowledge-survey.ANOVA AJAX                     -
   46 answers from   46 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  1
  None     21
  Some     24

Prob-knowledge-survey.bayes AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  3
  None     35
  Some      7

Prob-knowledge-survey.boot AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores      
        0
  None 45

Prob-knowledge-survey.bwplot AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12          6
  4-6           10
  6-9           18
  AP\n   course  3
  College        1
  Never         10

Prob-knowledge-survey.calc AJAX                     -
   47 answers from   47 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         14
  4-6            2
  6-9           23
  AP\n   course  3
  College        1
  Never          4

Prob-knowledge-survey.coll AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores      
        0
  None 34
  Some 11

Prob-knowledge-survey.cond AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores      
        0
  None 39
  Some  6

Prob-knowledge-survey.confI AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  3
  None     26
  Some     16

Prob-knowledge-survey.cprob AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         15
  4-6            1
  6-9            7
  AP\n   course  2
  College        1
  Never         22

Prob-knowledge-survey.cumu AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  1
  None     33
  Some     11

Prob-knowledge-survey.DF  AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  2
  None     25
  Some     18

Prob-knowledge-survey.dummy AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  1
  None     34
  Some     10

Prob-knowledge-survey.excel AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         16
  4-6            1
  6-9           11
  AP\n   course  2
  College        9
  Never          9

Prob-knowledge-survey.exp AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 15
  None      8
  Some     22

Prob-knowledge-survey.Ftest AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores      
        0
  None 45

Prob-knowledge-survey.hist AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores               
                 0
  1-3            2
  10-12         10
  4-6           10
  6-9           16
  AP\n   course  1
  College        1
  Never          8

Prob-knowledge-survey.Hnull AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 14
  None     14
  Some     17

Prob-knowledge-survey.intrct AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  1
  None     35
  Some      9

Prob-knowledge-survey.LA  AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         11
  4-6            4
  6-9            9
  AP\n   course  1
  College        5
  Never         18

Prob-knowledge-survey.lincomb AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  6
  None     26
  Some     13

Prob-knowledge-survey.logist AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores      
        0
  None 35
  Some 10

Prob-knowledge-survey.lreg AJAX                     -
   46 answers from   46 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 14
  None     14
  Some     18

Prob-knowledge-survey.LSQ AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  7
  None     26
  Some     12

Prob-knowledge-survey.main AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores      
        0
  None 34
  Some 11

Prob-knowledge-survey.mmm AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores         
           0
  1-3      6
  10-12    1
  4-6     26
  6-9      9
  College  1
  Never    2

Prob-knowledge-survey.mTerm AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores      
        0
  None 37
  Some  8

Prob-knowledge-survey.normal AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         23
  6-9           11
  AP\n   course  4
  College        5
  Never          5

Prob-knowledge-survey.ortho AJAX                     -
   44 answers from   44 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  3
  None     30
  Some     11

Prob-knowledge-survey.out AJAX                     -
   46 answers from   46 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 34
  None      2
  Some     10

Prob-knowledge-survey.perc AJAX                     -
   46 answers from   46 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 28
  None      1
  Some     17

Prob-knowledge-survey.prob AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores               
                 0
  1-3            4
  10-12         18
  4-6           12
  6-9           11
  AP\n   course  1
  College        1
  Never          1

Prob-knowledge-survey.program AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores               
                 0
  6-9            1
  AP\n   course  1
  College       15
  Never         31

Prob-knowledge-survey.pVal AJAX                     -
   46 answers from   46 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  9
  None     18
  Some     19

Prob-knowledge-survey.rand AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete 11
  None     17
  Some     17

Prob-knowledge-survey.rank AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores      
        0
  None 43
  Some  2

Prob-knowledge-survey.rcoef AJAX                     -
   46 answers from   46 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  8
  None     14
  Some     24

Prob-knowledge-survey.regres AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         17
  6-9            6
  AP\n   course  3
  College       10
  Never         12

Prob-knowledge-survey.resid1 AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  5
  None     18
  Some     22

Prob-knowledge-survey.sampD AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  3
  None     19
  Some     23

Prob-knowledge-survey.scatter AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores               
                 0
  10-12         11
  4-6           13
  6-9           18
  AP\n   course  2
  College        3
  Never          1

Prob-knowledge-survey.sd  AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores          
            0
  10-12     2
  6-9       2
  College   2
  Complete 21
  None      4
  Some     17

Prob-knowledge-survey.sig AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  4
  None     28
  Some     13

Prob-knowledge-survey.skew AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  3
  None     25
  Some     17

Prob-knowledge-survey.sResid AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  5
  None     29
  Some     11

Prob-knowledge-survey.subsp AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  5
  None     22
  Some     18

Prob-knowledge-survey.t   AJAX                     -
   48 answers from   48 readers. Mean score 0.00
Answers vs Scores               
                 0
  1-3            1
  10-12          8
  6-9            1
  AP\n   course  6
  College       10
  Never         22

Prob-knowledge-survey.tTest AJAX                     -
   46 answers from   46 readers. Mean score 0.00
Answers vs Scores          
            0
  Complete  3
  None     22
  Some     21

Prob-knowledge-survey.TypeI AJAX                     -
   45 answers from   45 readers. Mean score 0.00
Answers vs Scores      
        0
  None 36
  Some  9
</pre>
