Stats 155 Class Notes 2012-10-12
================================

```{r name="setup", child="notes-setup.Rmd"}
# boilerplate
```
### Hold Constant or Not

The in-class activity: 

[Total-vs-partial In-class activity](https://dl.dropbox.com/u/5098197/ISM/total-partial.pdf)



### Review

We saw a general strategy for modeling when the concern is with effect size:

1. Decide which sort of effect you want to study: total or partial.
2. Construct a model that includes both the variable you are interested in and any covariates that you want to hold constant.  What sort of model 
terms (e.g. interactions) you want to include is another matter, you have a choice.
3. Compute a partial derivative with respect to the variable whose effect size you want to study holding the other variables constant.

IMPORTANT: Leaving a covariate **out** of a model does not hold it constant.  Ignoring is not the same thing as holding constant.

### More than one variable?

It can occasionally happen that you want to study the effect of a variable that necessarily has an effect on another variable.  For instance, suppose we want to study the effect of education on wages.  But each year of education necessarily leads to a year less of "experience."  So, include both in the model:
```{r tidy=FALSE}
mod = lm( wage ~ sex*poly(exper,2)*educ + sector*educ, data=CPS85)
f = makeFun(mod)
f(sex="M", exper=9, educ=15,sector="prof") - f( sex="M", exper=10, educ=14, sector="prof")
```

You could also do this by constructing the differential involving both experience and education, but this will do.




