Power and Odds Ratios
========================================================

Consider a study with $n$ cases and $n$ controls, and an odds ratio of $R$.  Assuming that the two odds in the ratio are themselves reciprocals, with a value near $\sqrt{R}$, then it's as if the cases and controls are each drawn from a distribution with $n/(1+1/\sqrt{R})$ exposed cases with the remaining cases being unexposed.  Exposure is reversed for the controls.

```{r message=FALSE,error=FALSE,warning=FALSE,echo=FALSE}
require(mosaic)
```

```{r}
make.pop <- function(OR=2, popsize=10000) {
 cfrac = 1/(1+sqrt(OR))
 crep = round(10000*cfrac)
 arep = 10000 - crep
 cases= data.frame( exposed=c(rep("Exposed",arep),rep("Unexposed",crep)),
                    case = rep("Case",popsize))
 controls = data.frame( exposed=c(rep("Exposed",crep),rep("Unexposed",arep)),
                        case = rep("Control",popsize))
 pop = list( cases=cases, controls=controls )
 return(pop)
}
```

Standard odds ratio calculation with 95% CI
```{r}
OR.CI <- function(twobytwo) {
  a = twobytwo[1,1]
  b = twobytwo[1,2]
  c = twobytwo[2,1]
  d = twobytwo[2,2]
  OR = a*d/(c*b)
  se = sqrt( 1/a + 1/b + 1/c + 1/d)
  ci = exp( log(OR) + 2*se*c(-1,1))
  return( list(OR=OR, lower=ci[1], upper=ci[2]))
}
```
Do a simulation of an odds ratio and it's CI:

```{r}
OR.simulate <- function(OR, ncases=10, ncontrols=ncases,logscale=FALSE) {
  pop <- make.pop(OR=OR)
  if (logscale) logscale="x"
  else logscale=""
  plot(1:2, xlim=c(0.05,5*OR),ylim=c(0,100), type="n",log=logscale,
       xlab="Odds Ratio", ylab="Trial Number")
  lines( c(1,1),c(0,100),col="green")
  failedcount=0
  for (k in 1:100) {
    mydata <- rbind( sample(pop$cases, size=ncases), 
      sample(pop$controls, size=ncontrols) )
    mytab = tally( ~ exposed&case, data=mydata)
    res = OR.CI(mytab)
    color = "red"
    if (res$lower > 1 || res$upper < 1 ) { 
      color="blue"
      failedcount = failedcount + 1
    }
    lines( c(res$lower, res$upper), c(k,k), col=color )
    points( res$OR, k, col=color, pch=20)
  }
}
```

An example
```{r}    
OR.simulate(OR=3,ncases=50)
```
}
